---
title: "Cover Source Comparison: RAP vs Modelled Cover"
author: "Martin Holdrege"
date: "`r Sys.Date()`"
params:
  model_type: "herb"
  data_version: "d05"
  purer_version: "p04"
  model_version: "m12"
  cover_source: "glm_v01"
  n_sample: 50000
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 10, fig.height = 6)
is_woody <- params$model_type == "woody"
fig_height_pft <- if (is_woody) 10 else 6
```

## Purpose

Compare biomass predictions using RAP cover vs modelled (GLM) cover as the
cover multiplier. The model coefficients are identical — only the cover
input differs. This helps assess the impact of using modelled cover
(as will be done in SOILWAT2 projections) vs observed RAP cover.

## Setup

```{r source-and-load}
source("Functions/init.R")
library(patchwork)
source_functions()
theme_set(theme_bw())

model_type <- params$model_type
vd <- params$data_version
vp <- params$purer_version
vm <- params$model_version
cover_source <- params$cover_source

suffix <- paste0(model_type, "_", paste(vd, vp, vm, sep = "-"))
```

```{r load-model}
# --- load fitted model ---
p_fit <- file.path(
  paths$large,
  "Data_processed/BiomassQuantityData/Fit",
  paste0("fitted_model_", suffix, ".rds")
)
stopifnot(file.exists(p_fit))
m <- readRDS(p_fit)
fit <- m$fit
config <- m$config
cover_cols <- config$model$cover_cols
pred_vars <- config$model$pred_vars
```

```{r load-rasters}
# --- load CONUS rasters ---
rasters <- load_conus_rasters()
scale_df <- rasters$scale_df
# RAP cover
r_cover_rap <- rasters$cover[cover_cols]

# modelled cover
r_cover_mod <- load_modeled_cover(cover_source = cover_source,
                                   cover_cols = cover_cols)

# standardize climate
r_climate_std <- standardize_raster(rasters$climate[[pred_vars]],
                                     scale_df, vars = pred_vars)
```

```{r load-rap-predictions}
# --- load pre-computed RAP predictions ---
r_pred_rap <- load_pred_raster(suffix)
```

```{r predict-modelled-cover}
# --- predict with modelled cover ---
r_pred_stack_mod <- c(r_climate_std, r_cover_mod)

r_total_mod <- predict_raster(fit, r_pred_stack_mod, type = "total")
r_weighted_mod <- predict_raster(fit, r_pred_stack_mod, type = "by_group")
r_potential_mod <- predict_raster(fit, r_pred_stack_mod, type = "potential")
names(r_potential_mod) <- paste0(names(r_potential_mod), "_potential")

# extract RAP prediction layers
r_total_rap <- r_pred_rap[["predicted_totalBio"]]

weighted_lyrs <- names(r_pred_rap)[!grepl("_potential|predicted_totalBio",
                                          names(r_pred_rap))]
r_weighted_rap <- if (length(weighted_lyrs) > 0) r_pred_rap[[weighted_lyrs]] else NULL

potential_lyrs <- names(r_pred_rap)[grepl("_potential", names(r_pred_rap))]
r_potential_rap <- if (length(potential_lyrs) > 0) r_pred_rap[[potential_lyrs]] else NULL
```

```{r sample-for-scatterplots}
# --- sample from rasters for scatterplots ---
set.seed(42)

# cover comparison stack
r_cov_stack <- c(r_cover_rap, r_cover_mod)
rap_names <- paste0(cover_cols, "_rap")
mod_names <- paste0(cover_cols, "_mod")
names(r_cov_stack) <- c(rap_names, mod_names)

# prediction comparison stack
r_pred_stack <- c(r_total_rap, r_total_mod)
names(r_pred_stack) <- c("total_rap", "total_mod")

# add per-group if available
if (!is.null(r_weighted_rap)) {
  r_w_rap <- r_weighted_rap
  r_w_mod <- r_weighted_mod
  names(r_w_rap) <- paste0(names(r_w_rap), "_rap")
  names(r_w_mod) <- paste0(names(r_w_mod), "_mod")
  r_pred_stack <- c(r_pred_stack, r_w_rap, r_w_mod)
}

r_all <- c(r_cov_stack, r_pred_stack)
df_sample <- terra::spatSample(r_all, size = params$n_sample,
                                method = "random", na.rm = TRUE, xy = FALSE)
```

```{r print-config}
cat("Model type:", model_type, "\n")
cat("Model version:", vm, "\n")
cat("Cover source:", cover_source, "\n")
cat("Formula:", deparse(config$model$formula), "\n")
cat("Cover cols:", paste(cover_cols, collapse = ", "), "\n")
```

## 1. Cover Comparison: RAP vs Modelled {.tabset}

```{r cover-comparison, fig.width=14, fig.height=6, results='asis'}
for (i in seq_along(cover_cols)) {
  col <- cover_cols[i]
  rap_col <- paste0(col, "_rap")
  mod_col <- paste0(col, "_mod")
  
  cat("\n###", col, "\n\n")
  
  # maps side by side
  r_rap_i <- r_cover_rap[[col]]
  r_mod_i <- r_cover_mod[[col]]
  r_diff_cov <- r_mod_i - r_rap_i
  names(r_diff_cov) <- "difference"
  
  cov_q99 <- max(
    terra::global(r_rap_i, fun = function(x) quantile(x, 0.99, na.rm = TRUE))[[1]],
    terra::global(r_mod_i, fun = function(x) quantile(x, 0.99, na.rm = TRUE))[[1]]
  )
  
  diff_cov_q99 <- terra::global(
    abs(r_diff_cov),
    fun = function(x) quantile(x, 0.99, na.rm = TRUE)
  )[[1]]
  
  g1 <- plot_map_conus(r_rap_i,
                       colorscale = colorscale_cover(name = "Cover (%)",
                                                      limits = c(0, cov_q99)),
                       title = paste(col, "— RAP"))
  g2 <- plot_map_conus(r_mod_i,
                       colorscale = colorscale_cover(name = "Cover (%)",
                                                      limits = c(0, cov_q99)),
                       title = paste(col, "— Modelled"))
  g_diff <- plot_map_conus(r_diff_cov,
                           colorscale = colorscale_diverging(
                             name = "Mod − RAP",
                             limits = c(-diff_cov_q99, diff_cov_q99)),
                           title = paste(col, "— Difference (Mod − RAP)"))
  
  print(g1 + g2)
  print(g_diff)
  cat("\n\n")
  
  # scatterplot
  g_scatter <- ggplot(df_sample, aes(x = .data[[rap_col]],
                                      y = .data[[mod_col]])) +
    geom_hex(bins = 80) +
    scale_fill_viridis_c(trans = "log10") +
    geom_abline(slope = 1, intercept = 0, color = "red", linetype = "dashed") +
    labs(x = "RAP cover", y = "Modelled cover",
         title = paste(col, "— RAP vs Modelled")) +
    coord_equal()
  print(g_scatter)
  cat("\n\n")
}
```

## 2. Total Biomass Predictions

### Maps

```{r total-maps, fig.width=14, fig.height=6}
total_q99 <- max(
  terra::global(r_total_rap, fun = function(x) quantile(x, 0.99, na.rm = TRUE))[[1]],
  terra::global(r_total_mod, fun = function(x) quantile(x, 0.99, na.rm = TRUE))[[1]]
)
total_limits <- c(0, total_q99)

g1 <- plot_map_conus(r_total_rap,
                     colorscale = colorscale_biomass(limits = total_limits),
                     title = "Total biomass — RAP cover")
g2 <- plot_map_conus(r_total_mod,
                     colorscale = colorscale_biomass(limits = total_limits),
                     title = "Total biomass — Modelled cover")
g1+g2
```

### Difference (Modelled − RAP cover)

```{r total-diff-map, fig.width=10, fig.height=6}
r_diff_total <- r_total_mod - r_total_rap
names(r_diff_total) <- "difference"

diff_q99 <- terra::global(
  abs(r_diff_total),
  fun = function(x) quantile(x, 0.99, na.rm = TRUE)
)[[1]]

plot_map_conus(
  r_diff_total,
  colorscale = colorscale_diverging(
    name = "Mod − RAP\n(Mg/ha)",
    limits = c(-diff_q99, diff_q99)
  ),
  title = "Difference in total biomass (Modelled − RAP cover)"
)
```

### Scatterplot

```{r total-scatter, fig.width=6, fig.height=6}
ggplot(df_sample, aes(x = total_rap, y = total_mod)) +
  geom_hex(bins = 100) +
  scale_fill_viridis_c(trans = "log10") +
  geom_abline(slope = 1, intercept = 0, color = "red", linetype = "dashed") +
  labs(x = "Predicted biomass (RAP cover)",
       y = "Predicted biomass (Modelled cover)",
       title = "Total biomass: RAP vs Modelled cover") +
  coord_equal()
```

## 3. Per-PFT Biomass: Cover-Weighted {.tabset}

```{r pft-weighted-cover, fig.width=14, fig.height=6, eval=!is.null(r_weighted_rap), results='asis'}
for (lyr in weighted_lyrs) {
  r_rap_lyr <- r_weighted_rap[[lyr]]
  r_mod_lyr <- r_weighted_mod[[lyr]]
  r_diff_lyr <- r_mod_lyr - r_rap_lyr
  
  lyr_q99 <- max(
    terra::global(r_rap_lyr, fun = function(x) quantile(x, 0.99, na.rm = TRUE))[[1]],
    terra::global(r_mod_lyr, fun = function(x) quantile(x, 0.99, na.rm = TRUE))[[1]]
  )
  
  diff_lyr_q99 <- terra::global(
    abs(r_diff_lyr),
    fun = function(x) quantile(x, 0.99, na.rm = TRUE)
  )[[1]]
  
  cat("\n###", lyr, "\n\n")
  
  g1 <- plot_map_conus(r_rap_lyr,
                       colorscale = colorscale_biomass(limits = c(0, lyr_q99)),
                       title = paste(lyr, "— RAP cover"))
  g2 <- plot_map_conus(r_mod_lyr,
                       colorscale = colorscale_biomass(limits = c(0, lyr_q99)),
                       title = paste(lyr, "— Modelled cover"))
  g_diff <- plot_map_conus(r_diff_lyr,
                           colorscale = colorscale_diverging(
                             name = "Mod-RAP",
                             limits = c(-diff_lyr_q99, diff_lyr_q99)),
                           title = paste(lyr, "— Difference"))
  
  print(g1)
  cat("\n\n")
  print(g2)
  cat("\n\n")
  print(g_diff)
  cat("\n\n")
  
  # scatterplot
  rap_col <- paste0(lyr, "_rap")
  mod_col <- paste0(lyr, "_mod")
  if (all(c(rap_col, mod_col) %in% names(df_sample))) {
    g_scatter <- ggplot(df_sample, aes(x = .data[[rap_col]],
                                        y = .data[[mod_col]])) +
      geom_hex(bins = 80) +
      scale_fill_viridis_c(trans = "log10") +
      geom_abline(slope = 1, intercept = 0, color = "red", linetype = "dashed") +
      labs(x = paste(lyr, "(RAP)"), y = paste(lyr, "(Modelled)"),
           title = paste(lyr, "— RAP vs Modelled cover")) +
      coord_equal()
    print(g_scatter)
    cat("\n\n")
  }
}
```

## 4. Per-PFT Biomass: Potential (Cover = 100%)

Potential biomass is identical between RAP and modelled cover since it
uses cover = 100%. This section confirms that — any differences indicate a bug.

```{r potential-check, fig.width=6, fig.height=6, eval=!is.null(r_potential_rap)}
# potential should be identical since cover is not used
r_diff_pot <- r_potential_mod - r_potential_rap
max_diff <- terra::global(abs(r_diff_pot), "max", na.rm = TRUE)
cat("Max absolute difference in potential biomass:", max(max_diff$max), "\n")
cat("(Should be ~0 — potential biomass doesn't depend on cover source)\n")
```

## Session Info

```{r session-info}
cat("Model:", suffix, "\n")
cat("Cover source:", cover_source, "\n")
cat("N sampled:", params$n_sample, "\n")
cat("Rendered:", format(Sys.time(), "%Y-%m-%d %H:%M:%S"), "\n")
```
