---
title: "Model Evaluation: Simulated Data"
author: "Martin Holdrege"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.width = 7, fig.height = 5)
```

## Purpose

Evaluate the fitted cwexp model against known simulation truth. This script
is specific to simulated data where we know the true parameters and per-PFT
biomass.

## Setup


```{r source-and-load}
source('Functions/init.R')
source_functions()
library(ggplot2)
library(dplyr)
theme_set(theme_bw())

# --- load fitted model output ---
vp <- 'p01' # version of purer selection
vm <- 'm01' # version of model
suffix <- paste(opt$vs, vp, vm, sep = '-')
test_run <- FALSE
if(test_run) suffix <- 'test_run'
# created in
# "Analysis/BiomassQuantity/Fit/01_fit_model.R"
p_fit <- file.path(
  paths$large,
  'Data_processed/BiomassQuantityData/Fit',
  paste0('fitted_model_', suffix, '.rds')
)
m1 <- readRDS(p_fit)

# --- load simulation truth ---
p_truth <- file.path(
  paths$large,
  'Data_processed/BiomassQuantityData/simulated',
  paste0('simBiomass_', opt$vs, '.rds')
)
truth <- readRDS(p_truth)

# --- evaluation data (full simulated dataset) ---
dat_eval <- truth$data

# --- extract key objects ---
fit <- m1$fit
cv <- m1$cv
config <- m1$config
```

```{r print-config}
cat("Model version:", vm, "\n")
cat("Data version:", opt$vs, "\n")
cat("Formula:", deparse(config$model$formula), "\n")
cat("Selected lambda:", cv$selected$lambda, "\n")
cat("Selection rule:", config$cv$select_rule, "\n")
cat("N training (purer pixels):", m1$purer$n_selected, "\n")
cat("N evaluation (full data):", nrow(dat_eval), "\n")
```

## 1. Overall Mu Recovery

```{r mu-scatter, fig.width=7, fig.height=4.5}
mu_true <- dat_eval$totalMu
mu_hat_reg <- predict(fit, dat_eval, type = "mu")
hist(mu_hat_reg)
hist(mu_true)

lims <- quantile(c(mu_hat_reg, mu_true), c(0, 0.9))
plot(mu_true, mu_hat_reg, pch = 16, cex = 0.3, col = rgb(0, 0, 0, 0.2),
     xlab = "True mu", ylab = "Predicted mu",
     xlim = lims, ylim = lims,
     main = paste0("Regularized (lambda = ", round(cv$selected$lambda, 4), ")"))
abline(0, 1, col = "red", lwd = 2)
par(mfrow = c(1, 1))
```

## 2. B Matrix Recovery

```{r B-recovery}
aligned <- align_B_matrices(truth$par$B, fit$par$B)

if (length(aligned$extra_in_fit) > 0) {
  cat("Predictors in fitted model but not in truth (compared to 0):\n ",
      paste(aligned$extra_in_fit, collapse = ", "), "\n")
}
if (length(aligned$missing_in_fit) > 0) {
  cat("Predictors in truth but not in fitted model (excluded):\n ",
      paste(aligned$missing_in_fit, collapse = ", "), "\n")
}
```

```{r B-scatter, fig.width=10, fig.height=4.5}


B_true_vec <- as.numeric(aligned$true)
B_reg_vec <- as.numeric(aligned$fit)

# tag extra predictors (truth = 0)
extra_rows <- rownames(aligned$true) %in% aligned$extra_in_fit
is_extra <- rep(extra_rows, ncol(aligned$true))


plot(B_true_vec, B_reg_vec, pch = 16,
     col = ifelse(is_extra, "red", "black"),
     xlab = "True B", ylab = "Fitted B",
     main = "B recovery: regularized")
abline(0, 1, col = "grey50", lty = 2)
abline(h = 0, v = 0, col = "grey80")

if (any(is_extra)) {
  legend("topleft", legend = c("Shared predictors", "Extra in fit (true = 0)"),
         col = c("black", "red"), pch = 16, bty = "n")
}
```

## 3. Alpha Recovery

```{r alpha-recovery}
# align by name
shared_groups <- intersect(names(fit$par$alpha), names(truth$par$alpha))

alpha_df <- tibble(
  group = shared_groups,
  true_alpha = truth$par$alpha[shared_groups],
  reg_alpha = fit$par$alpha[shared_groups]
) |>
  mutate(across(where(is.numeric), \(x) round(x, 2)))

```

```{r alpha-scatter, fig.width=5, fig.height=4.5}
plot(alpha_df$true_alpha, alpha_df$reg_alpha, pch = 16, cex = 1.5,
     xlab = "True alpha", ylab = "Fitted alpha",
     main = "Alpha recovery (regularized)")
abline(0, 1)
```

## 4. Per-Group Recovery

```{r group-table}
compare_group_to_truth(
  truth = truth,
  fits = list(regularized = fit),
  newdata = dat_eval,
  use_observed = 'force'
) |>
  mutate(across(where(is.numeric), \(x) round(x, 3))) |>
  knitr::kable(caption = "Per-group mu recovery on full dataset")
```

```{r group-plots, fig.width=8, fig.height=6}

plot_group_recovery(truth = truth, fit = fit, 
                    newdata = dat_eval,
                    density = TRUE,
                    use_observed = 'force', 
                    title = "Per-PFT: regularized") +
  geom_smooth(se = FALSE)


plot_group_recovery(truth = truth, fit = fit, 
                    newdata = dat_eval,
                    weighted = FALSE,
                    density = TRUE,
                    use_observed = 'force', 
                    title = "Per-PFT: regularized, (biomass at 100% cover)") +
  geom_smooth(se = FALSE)
```

## 5. CV Diagnostics: Score vs Lambda

```{r cv-score-lambda, fig.width=7, fig.height=4}
score_summary <- cv$score_summary

ggplot(score_summary, aes(x = lambda, y = mae_log)) +
  geom_line() +
  geom_point() +
  geom_errorbar(aes(ymin = mae_log - mae_log_se,
                    ymax = mae_log + mae_log_se),
                width = 0) +
  geom_vline(xintercept = cv$selected$lambda,
             linetype = "dashed", color = "red") +
  labs(x = "Lambda", y = "MAE (log scale)",
       title = "CV holdout score vs lambda",
       subtitle = paste("Selected lambda =",
                        signif(cv$selected$lambda, 2),
                        "(rule:", config$cv$select_rule, ")"))
```

## 6. Truth Recovery vs Lambda

Using the stored per-fold parameters at each lambda to compute truth metrics
across the full lambda path.

```{r truth-vs-lambda}
# helper: wrap stored par into a minimal object predict() can use
make_par_obj <- function(par_list, spec) {
  obj <- list(
    par = list(
      alpha = par_list$alpha,
      B = par_list$B,
      sigma = if ("sigma" %in% names(par_list)) par_list$sigma else exp(par_list$log_sigma)
    ),
    spec = spec
  )
  class(obj) <- "cwexp_tmb_fit"
  obj
}

# compute truth metrics at each lambda (averaged across folds)
path_par <- cv$path_par
spec <- cv$spec

truth_by_lambda <- path_par |>
  group_by(lambda) |>
  summarise(
    mu_rmse = mean(purrr::map_dbl(par, function(p) {
      obj <- make_par_obj(p, spec)
      mu_hat <- predict(obj, dat_eval, type = "mu")
      metric_rmse(mu_true, mu_hat)
    })),
    B_rmse = mean(purrr::map_dbl(par, function(p) {
      aligned_i <- align_B_matrices(truth$par$B,
                                     {
                                       B_i <- p$B
                                       rownames(B_i) <- spec$x_cols %||% rownames(truth$par$B)
                                       colnames(B_i) <- spec$cover_cols
                                       B_i
                                     })
      metric_rmse(as.numeric(aligned_i$true), as.numeric(aligned_i$fit))
    })),
    .groups = "drop"
  )
```

```{r truth-vs-lambda-plot, fig.width=10, fig.height=4}
par(mfrow = c(1, 2))

plot(truth_by_lambda$lambda, truth_by_lambda$mu_rmse,
     type = "b", pch = 19,
     xlab = "Lambda", ylab = "rmse(true mu, predicted mu)",
     main = "Mu recovery vs lambda")
abline(v = cv$selected$lambda, col = "red", lty = 2)

plot(truth_by_lambda$lambda, truth_by_lambda$B_rmse,
     type = "b", pch = 19,
     xlab = "Lambda", ylab = "rmse(true B, fitted B)",
     main = "B recovery vs lambda")
abline(v = cv$selected$lambda, col = "red", lty = 2)

par(mfrow = c(1, 1))
```

## Summary

```{r summary}
cat("Model:", deparse(config$model$formula), "\n")
cat("Selected lambda:", signif(cv$selected$lambda, 2),
    "(", config$cv$select_rule, "rule )\n")
cat("Mu cor (regularized, full data):", round(cor(mu_true, mu_hat_reg), 3), "\n")
cat("B cor (regularized):", round(cor(B_true_vec, B_reg_vec), 3), "\n")
```
