---
title: "Model Diagnostics"
author: "Martin Holdrege"
date: "`r Sys.Date()`"
params:
  model_type: "woody"
  data_version: "d05"
  purer_version: 'p04'
  model_version: "m12"
  quantile_vars: !r c("MAT", "MAP", "PrecipTempCorr", "awc", "isothermality")
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE,
                      fig.width = 8, fig.height = 5)
```

## Purpose

Model diagnostics for the cwexp biomass model. Parameterized by model type
(`herb` or `woody`), data version, and model version.

## Setup

```{r source-and-load}
source("Functions/init.R")
library(patchwork)
source_functions()
theme_set(theme_bw())

vm <- params$model_version
vd <- params$data_version
vp <- params$purer_version
model_type <- params$model_type
is_woody <- model_type == 'woody'
# --- load fitted model ---
suffix <- paste0(model_type, "_", paste(vd, vp, vm, sep = "-"))

m1 <- load_fit(suffix)

fit <- m1$fit
cv <- m1$cv
config <- m1$config
n_pft <- length(m1$fit$spec$cover_cols)
# --- load full dataset ---

data_obj <- read_analysis_ready_hw(vd = vd, model_type = model_type,
                                   only_dataframe = FALSE)

dat_full <- data_obj$data
if(nrow(dat_full) > 5e5) {
  
  dat_full <- dat_full[sample.int(nrow(dat_full), size = 5e5), ]
}
scale_df <- data_obj$scale

# --- predictions on full data ---
dat_full$predicted <- predict(fit, dat_full, type = "mu")

dat_full$observed <- dat_full$totalBio
set.seed(123)
```

```{r}
# CV predictions (i.e. each cv model used to predict on held 
# out environmental clusters)
dat_full$predicted_cv <- cv_holdout_predictions(m1, data = dat_full)
```


```{r unstandardize}
dat_full_orig <- unstandardize(dat_full, scale_df)
```

```{r print-config}
cat("Model type:", model_type, "\n")
cat("Model version:", vm, "\n")
cat("Data version:", vd, "\n")
cat("Purer version:", vp, "\n")
cat("Formula:", deparse(config$model$formula), "\n")
cat("Cover variables:", paste(config$model$cover_cols, collapse = ", "), "\n")
cat("EN alpha:", config$cv$en_alpha, "\n")
cat("Selected lambda:", signif(cv$selected$lambda, 3), "\n")
cat("Selection rule:", config$cv$select_rule, "\n")
cat("N training:", m1$n_train, "\n")
cat("N full data:", nrow(dat_full), "\n")
```


## 1. Lambda Selection

Cross-validation score across the lambda path, with the selected lambda
marked.

```{r cv-score-plots, fig.width=10, fig.height=7}
plot_cv_scores(score_summary = cv$score_summary, 
               selected_lambda = cv$selected$lambda)
plot_cv_scores(score_summary = cv$score_summary, 
               selected_lambda = cv$selected$lambda,
               x_scale = 'rank')
```

```{r selected-lambda-info}
cat("Selected lambda:", signif(cv$selected$lambda, 3), "\n")
cat("Rule:", config$cv$select_rule, "\n\n")

cat("Scores at selected lambda:\n")
cv$selected |>
  mutate(across(where(is.numeric), \(x) round(x, 3))) |>
  as.data.frame() |>
  print()
```

## 2. CV Holdout Error on Total Biomass

Per-fold holdout prediction errors from cross-validation. These are honest
out-of-sample metrics since each fold's test data was not used for training
or lambda selection.

```{r cv-fold-scores}
# scores at the selected lambda only
fold_scores_at_selected <- cv$scores |>
  filter(abs(lambda - cv$selected$lambda) < 1e-10)

```

```{r cv-fold-summary}
# summary across folds
fold_summary <- fold_scores_at_selected |>
  summarise(
    across(c(mae_log1p, rmse_log1p, mae, rmse, cor),
           list(mean = mean, sd = sd),
           .names = "{.col}_{.fn}")
  ) |>
  pivot_longer(everything(),
               names_to = c("metric", "stat"),
               names_sep = "_(?=[^_]+$)") |>
  pivot_wider(names_from = stat, values_from = value) |>
  mutate(across(where(is.numeric), \(x) round(x, 4)))

knitr::kable(fold_summary,
             caption = "CV holdout error summary (mean ± sd across folds)")
```

## 3. Final Model Error on Full Dataset

Predictions from the global model (fitted at the selected lambda on the
purer training data) evaluated on the full dataset. This is optimistic
since it includes the training data, but shows overall model behavior.

```{r final-model-metrics}
metrics_df <- cv_metrics_df(dat_full) |> 
  mutate(value = signif(value, 3))

knitr::kable(metrics_df, caption = "Final model error on full dataset")
```

```{r obs-vs-pred-scatter, fig.width=10, fig.height=4.5}
plot_obs_vs_pred(dat_full, title = 'Observed vs predicted total biomass',
                 x_percentile = 0.99, y_percentile = 0.99)
plot_obs_vs_pred(dat_full, predicted = 'predicted_cv', 
                 title = 'Observed vs predicted total biomass (CV)',
                 x_percentile = 0.99,y_percentile = 0.99)
```

```{r residual-plots, fig.width=10, fig.height=4.5}
plot_residual(dat_full)

```


```{r residual-ll, fig.width=10, fig.height=4.5}
plot_loglik(dat_full, fit, log1p_x = FALSE, size = 1e5)
plot_loglik(dat_full, fit, log1p_x = TRUE, size = 1e5)

```

What is the bias in mean predictions on the original data scale?

```{r mean-bias}
mean(dat_full$totalBio) / mean(dat_full$predicted)

plot_bias_ratio(dat_full)
```



## 4. 'Holdrege' plots: Total Biomass Across Climate 

Mean observed and predicted total biomass within percentile bins of each
predictor variable.

```{r quantile-setup}
quantile_vars <- params$quantile_vars

# decile functions expect yvar and yvar_pred naming convention
dat_quant <- dat_full_orig |>
  dplyr::rename(totalBio_pred = predicted)
```


```{r quantile-total-bio, fig.width=10, fig.height=7}

fns <- c('mean', 'median')
quant_dfs <- list()
for (f in fns) {
  quant_df <- predvars2deciles(
    df = dat_quant,
    response_vars = c("totalBio", "totalBio_pred", 'predicted_cv'),
    pred_vars = quantile_vars,
    summary_fun = f
  )
  quant_dfs[[f]] <- quant_df
  g <- decile_dotplot(
    yvar = "totalBio",
    df = quant_df,
    add_predicted = TRUE,
    ylab = paste0("Total Biomass (", f, ")"),
    xlab = paste(f, "of quantile of predictor variable"),
    title = "Observed vs Predicted totalBio across climate percentiles"
  )
  print(g)
}

quant_df <- quant_dfs$mean
```

```{r quantile-filtered, fig.width=12, fig.height=7}

f <- 'mean'
filter_vars <- c('MAT', 'MAP', 'PrecipTempCorr')

quant_df_filt <- predvars2deciles(
  df = dat_quant,
  response_vars = c("totalBio", "totalBio_pred"),
  pred_vars = quantile_vars,
  filter_var = TRUE,
  filter_vars = filter_vars,
  summary_fun = f
)

g <- decile_dotplot_filtered(
  yvar = "totalBio",
  df = quant_df_filt,
  ylab = paste0("Total Biomass (", f, ")"),
  xlab = paste(f, "of quantile of predictor variable"),
  title = "Observed vs Predicted totalBio across climate percentiles"
)
print(g)


```

### CV based holdrege plots

Predictions made by each CV model to the held out folds. 


```{r quantile-cv, fig.width=10, fig.height=7}

quant_df_cv <- quant_df |> 
  select(-totalBio_pred) |> 
  rename(totalBio_pred = predicted_cv)
g <- decile_dotplot(
  yvar = "totalBio",
  df = quant_df_cv,
  add_predicted = TRUE,
  ylab = paste0("Total Biomass (mean)"),
  title = "Observed vs Predicted totalBio (CV predictions)"
)
print(g)


```


## 5. Holdrege plots: Per-PFT Biomass Across Climate


### Cover-Weighted

```{r pft-weighted, fig.width=12, fig.height=n_pft*4}

quant_weighted <- pft_quantile_summary(
  fit = fit, data = dat_full,
  formula = config$model$formula,
  cover_cols = config$model$cover_cols,
  pred_vars = params$quantile_vars,
  weighted = TRUE,
  summary_fun = 'median'
)

quant_weighted_orig <- unstandardize_long(data = quant_weighted, scale_df = scale_df,
                                          name_col = 'name', value_col = 'mean_value')
  
plot_pft_quantiles(quant_weighted_orig,
                   title = "Per-PFT biomass across climate (cover-weighted)")

```

### Potential Biomass (Cover = 100%)

```{r pft-potential, fig.width=12, fig.height=n_pft*4}
quant_potential <- pft_quantile_summary(
  fit = fit, data = dat_full,
  formula = config$model$formula,
  cover_cols = config$model$cover_cols,
  pred_vars = params$quantile_vars,
  weighted = FALSE
) |> 
  unstandardize_long(scale_df = scale_df)

plot_pft_quantiles(quant_potential,
                   title = "Per-PFT potential biomass across climate (if cover = 100%)",
                   ylab = "Potential biomass (if cover = 100%)")

```

## Cofficients

```{r}
plot_alpha_vs_lambda(cv$path_par)
```

```{r coefs}
lapply(fit$par, round, 2)
```

## 6. Partial Dependence Plots

### Total Biomass

```{r pdp-total, fig.width=6, fig.height=4}
pdp_total <- cwexp_pdp(
  fit = fit, data = dat_full,
  type = "total"
) |> 
  unstandardize_long(scale_df, name_col = 'variable',
                     value_col = 'x_value')

plot_cwexp_pdp(pdp_total,
               title = "Partial dependence: total biomass")
```

### Per-PFT (Cover-Weighted)
```{r pdp-pft-weighted, fig.width=12, fig.height=n_pft*4}
pdp_weighted <- cwexp_pdp(
  fit = fit, data = dat_full,
  type = "by_group",
  weighted = TRUE
)|> 
  unstandardize_long(scale_df, name_col = 'variable',
                     value_col = 'x_value')

plot_cwexp_pdp(pdp_weighted,
               title = "Partial dependence: per-PFT biomass (cover-weighted)")
```

### Per-PFT (Potential, Cover = 100%)

```{r pdp-pft-potential, fig.width=12, fig.height=n_pft*4}
pdp_potential <- cwexp_pdp(
  fit = fit, data = dat_full,
  type = "by_group",
  weighted = FALSE
)|> 
  unstandardize_long(scale_df, name_col = 'variable',
                     value_col = 'x_value')

plot_cwexp_pdp(pdp_potential,
               title = "Partial dependence: per-PFT potential biomass (cover = 100%)",
               ylab = "Potential biomass (at 100% cover)")
```

```{asis, echo=is_woody}
## 7. Comparison with Field Shrub Biomass

Independent field measurements of shrub biomass from three studies
(Wright, Jornada LTER, Sevilleta LTER) compared against model predictions
extracted at the same locations.
```

```{r shrub-field-load, eval=is_woody}
shrub_obs <- read_shrub_biomass()

plot_points_conus(shrub_obs, color_var = 'study')
```

```{r shrub-histogram, fig.width=8, fig.height=4, eval=is_woody}

ggplot(shrub_obs, aes(shrubBiomass)) +
  geom_histogram() +
  facet_wrap(~study) +
  labs(x = "Observed shrub biomass (Mg/ha)", y = "Count")

ggplot(shrub_obs, aes(shrubBiomass_potential)) +
  geom_histogram() +
  facet_wrap(~study) +
  labs(x = "Observed potential shrub biomass (Mg/ha)", y = "Count")
```

```{r shrub-extract, eval=is_woody}

r_pred_all <- load_pred_raster(suffix) 

extracted <- terra::extract(
  r_pred_all[[c("totalShrub", "totalShrub_potential")]],
  terra::vect(shrub_obs)
)

shrub_compare <- bind_cols(shrub_obs, extracted) |>
  sf::st_drop_geometry() |>
  rename(pred_shrubBio = totalShrub,
         pred_shrubBio_potential = totalShrub_potential)
```

```{asis, echo=is_woody}
### Observed vs Predicted Shrub Biomass (Cover-Weighted)
```

```{r shrub-scatter-weighted, fig.width=6, fig.height=5, eval=is_woody}
ggplot(shrub_compare, aes(x = shrubBiomass, y = pred_shrubBio, color = study)) +
  geom_point(alpha = 0.6) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "grey40") +
  labs(x = "Observed shrub biomass (Mg/ha)",
       y = "Predicted shrub biomass (Mg/ha)",
       title = "Cover-weighted shrub biomass: observed vs predicted")
```

```{asis, echo=is_woody}
### Observed vs Predicted Shrub Biomass (Potential)
```


```{r shrub-scatter-potential, fig.width=6, fig.height=5, eval=is_woody}
ggplot(shrub_compare, aes(x = shrubBiomass_potential, y = pred_shrubBio_potential,
                          color = study)) +
  geom_point(alpha = 0.6) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "grey40") +
  labs(x = "Observed potential shrub biomass (Mg/ha)",
       y = "Predicted potential shrub biomass (Mg/ha)",
       title = "Potential shrub biomass: observed vs predicted")
```

```{asis, echo=is_woody}
### Summary Metrics
```

```{r shrub-metrics}
shrub_metrics <- shrub_compare |>
  summarise(
    n = n(),
    cor_weighted = cor(shrubBiomass, pred_shrubBio, use = "complete.obs"),
    rmse_weighted = sqrt(mean((shrubBiomass - pred_shrubBio)^2, na.rm = TRUE)),
    cor_potential = cor(shrubBiomass_potential, pred_shrubBio_potential, use = "complete.obs"),
    rmse_potential = sqrt(mean((shrubBiomass_potential - pred_shrubBio_potential)^2, na.rm = TRUE))
  ) |>
  mutate(across(where(is.numeric), \(x) signif(x, 3)))

knitr::kable(shrub_metrics,
             caption = "Field shrub biomass vs model predictions")
```

## Session Info

```{r session-info}
cat("Rendered:", format(Sys.time(), "%Y-%m-%d %H:%M:%S"), "\n")
```
