---
title: "Model Diagnostics"
author: "Martin Holdrege"
date: "`r Sys.Date()`"
params:
  model_version: "m04"
  data_version: "d02"
  purer_version: 'p02'
  # climate vars to bin by in quantile plots (sections 4-5)
  quantile_vars: !r c("tmean_CLIM", "precip_CLIM", "PrecipTempCorr_CLIM")
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.width = 8, fig.height = 5)
```

## Purpose

Model diagnostics for the cwexp biomass model. This report works for both
simulated and real data — it evaluates the fitted model without requiring
knowledge of the true parameters.

## Setup

```{r source-and-load}
source("Functions/init.R")
library(patchwork)
source_functions()
theme_set(theme_bw())

vm <- params$model_version
vd <- params$data_version
vp <- params$purer_version

# --- load fitted model ---
suffix <- paste(vd, vp, vm, sep = '-')

p_fit <- file.path(
  paths$large,
  "Data_processed/BiomassQuantityData/Fit",
  paste0("fitted_model_", suffix, ".rds")
)
stopifnot(isTRUE(file.exists(p_fit)))
m1 <- readRDS(p_fit)

fit <- m1$fit
cv <- m1$cv
config <- m1$config

# --- load full dataset ---
dat_full <- read_analysis_ready(vd = vd)

# --- predictions on full data ---
mu_hat <- predict(fit, dat_full, type = "mu")
dat_full$predicted <- mu_hat
dat_full$observed <- dat_full$totalBio
```

```{r print-config}
cat("Model version:", vm, "\n")
cat("Data version:", vd, "\n")
cat("Sampling (purer) version:", vp, '\n')
cat("Formula:", deparse(config$model$formula), "\n")
cat("EN alpha:", config$cv$en_alpha, "\n")
cat("Selected lambda:", round(cv$selected$lambda, 5), "\n")
cat("Selection rule:", config$cv$select_rule, "\n")
cat("N training (purer pixels):", m1$purer$n_selected, "\n")
cat("N full data:", nrow(dat_full), "\n")
```

## 1. Lambda Selection

Cross-validation score across the lambda path, with the selected lambda
marked.

```{r cv-score-plots, fig.width=10, fig.height=7}
plot_cv_scores(score_summary = cv$score_summary, 
               selected_lambda = cv$selected$lambda)
plot_cv_scores(score_summary = cv$score_summary, 
               selected_lambda = cv$selected$lambda,
               x_scale = 'rank')
```

```{r selected-lambda-info}
cat("Selected lambda:", signif(cv$selected$lambda, 3), "\n")
cat("Rule:", config$cv$select_rule, "\n\n")

cat("Scores at selected lambda:\n")
cv$selected |>
  mutate(across(where(is.numeric), \(x) round(x, 3))) |>
  as.data.frame() |>
  print()
```

## 2. CV Holdout Error on Total Biomass

Per-fold holdout prediction errors from cross-validation. These are honest
out-of-sample metrics since each fold's test data was not used for training
or lambda selection.

```{r cv-fold-scores}
# scores at the selected lambda only
fold_scores_at_selected <- cv$scores |>
  filter(abs(lambda - cv$selected$lambda) < 1e-10)

```

```{r cv-fold-summary}
# summary across folds
fold_summary <- fold_scores_at_selected |>
  summarise(
    across(c(mae_log, rmse_log, mae, rmse, cor),
           list(mean = mean, sd = sd),
           .names = "{.col}_{.fn}")
  ) |>
  pivot_longer(everything(),
               names_to = c("metric", "stat"),
               names_sep = "_(?=[^_]+$)") |>
  pivot_wider(names_from = stat, values_from = value) |>
  mutate(across(where(is.numeric), \(x) round(x, 4)))

knitr::kable(fold_summary,
             caption = "CV holdout error summary (mean ± sd across folds)")
```

## 3. Final Model Error on Full Dataset

Predictions from the global model (fitted at the selected lambda on the
purer training data) evaluated on the full dataset. This is optimistic
since it includes the training data, but shows overall model behavior.

```{r final-model-metrics}
metrics_df <- cv_metrics_df(dat_full) |> 
  mutate(value = signif(value, 3))

knitr::kable(metrics_df, caption = "Final model error on full dataset")
```

```{r obs-vs-pred-scatter, fig.width=10, fig.height=4.5}
plot_obs_vs_pred(dat_full, title = 'Observed vs predicted total biomass')
```

```{r residual-plots, fig.width=10, fig.height=4.5}
plot_residual(dat_full)

```

## 4. Total Biomass Across Climate (Quantile Plots)

Mean observed and predicted total biomass within percentile bins of each
predictor variable.

```{r quantile-setup}
quantile_vars <- params$quantile_vars

# decile functions expect yvar and yvar_pred naming convention
dat_quant <- dat_full |>
  dplyr::rename(totalBio_pred = predicted)
```


```{r quantile-total-bio, fig.width=10, fig.height=7}

for (f in c('mean', 'median')) {
  quant_df <- predvars2deciles(
    df = dat_quant,
    response_vars = c("totalBio", "totalBio_pred"),
    pred_vars = quantile_vars,
    summary_fun = f
  )

  g <- decile_dotplot(
    yvar = "totalBio",
    df = quant_df,
    add_predicted = TRUE,
    ylab = paste0("Total Biomass (", f, ")"),
    xlab = paste(f, "of quantile of predictor variable"),
    title = "Observed vs Predicted totalBio across climate percentiles"
  )
  print(g)
}

```


## 5. Per-PFT Biomass Across Climate


### Cover-Weighted
```{r pft-weighted, fig.width=12, fig.height=10}

quant_weighted <- pft_quantile_summary(
  fit = fit, data = dat_full,
  formula = config$model$formula,
  cover_cols = config$model$cover_cols,
  pred_vars = params$quantile_vars,
  weighted = TRUE,
  summary_fun = 'median'
)

plot_pft_quantiles(quant_weighted,
                   title = "Per-PFT biomass across climate (cover-weighted)")

```

### Potential Biomass (Cover = 100%)

```{r pft-potential, fig.width=12, fig.height=10}
quant_potential <- pft_quantile_summary(
  fit = fit, data = dat_full,
  formula = config$model$formula,
  cover_cols = config$model$cover_cols,
  pred_vars = params$quantile_vars,
  weighted = FALSE
)

plot_pft_quantiles(quant_potential,
                   title = "Per-PFT potential biomass across climate (if cover = 100%)",
                   ylab = "Potential biomass (if cover = 100%)")
```

## Cofficients

```{r}
plot_alpha_vs_lambda(cv$path_par)
```

```{r coefs}
lapply(fit$par, round, 2)
```

## 6. Partial Dependence Plots

### Total Biomass

```{r pdp-total, fig.width=6, fig.height=4}
pdp_total <- cwexp_pdp(
  fit = fit, data = dat_full,
  type = "total"
)

plot_cwexp_pdp(pdp_total,
               title = "Partial dependence: total biomass")
```

### Per-PFT (Cover-Weighted)
```{r pdp-pft-weighted, fig.width=12, fig.height=10}
pdp_weighted <- cwexp_pdp(
  fit = fit, data = dat_full,
  type = "by_group",
  weighted = TRUE
)

plot_cwexp_pdp(pdp_weighted,
               title = "Partial dependence: per-PFT biomass (cover-weighted)")
```

### Per-PFT (Potential, Cover = 100%)

```{r pdp-pft-potential, fig.width=12, fig.height=10}
pdp_potential <- cwexp_pdp(
  fit = fit, data = dat_full,
  type = "by_group",
  weighted = FALSE
)

plot_cwexp_pdp(pdp_potential,
               title = "Partial dependence: per-PFT potential biomass (cover = 100%)",
               ylab = "Potential biomass (at 100% cover)")
```
## Session Info

```{r session-info}
cat("Model file:", p_fit, "\n")
cat("Rendered:", format(Sys.time(), "%Y-%m-%d %H:%M:%S"), "\n")
```
