---
title: "Model Comparison"
author: "Martin Holdrege"
date: "`r Sys.Date()`"
params:
  model_type: "woody"
  # model 1 (reference)
  vd1: "d05"
  vp1: "p04"
  vm1: "m12"
  # model 2 (comparison)
  vd2: "d05"
  vp2: "p04"
  vm2: "m13"
  n_sample_raster: 50000
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 10, fig.height = 6)
is_woody <- params$model_type == "woody"
fig_height_pft <- if (is_woody) 10 else 6
```

## Purpose

Side-by-side comparison of two `r params$model_type` cwexp biomass models.
Compares predicted biomass maps, per-PFT decomposition, and scatterplots
of model 1 vs model 2 predictions sampled from the prediction rasters.

## Setup

```{r source-and-load}
source("Functions/init.R")
library(patchwork)
source_functions()
theme_set(theme_bw())

model_type <- params$model_type

# build suffixes
suffix1 <- paste0(model_type, "_", paste(params$vd1, params$vp1, params$vm1, sep = "-"))
suffix2 <- paste0(model_type, "_", paste(params$vd2, params$vp2, params$vm2, sep = "-"))

label1 <- paste(params$vd1, params$vp1, params$vm1, sep = "-")
label2 <- paste(params$vd2, params$vp2, params$vm2, sep = "-")
```

```{r load-models}
# --- load fitted models ---

m1 <- load_fit(suffix1)
m2 <- load_fit(suffix2)

# verify same PFT structure
stopifnot(identical(m1$config$model$cover_cols, m2$config$model$cover_cols))
cover_cols <- m1$config$model$cover_cols
n_pft <- length(cover_cols)
```


```{r load-prediction-rasters}
# --- load prediction rasters ---
r1 <- load_pred_raster(suffix1)
r2 <- load_pred_raster(suffix2)

cat("Model 1 layers:", paste(names(r1), collapse = ", "), "\n")
cat("Model 2 layers:", paste(names(r2), collapse = ", "), "\n")
```

```{r extract-layers}
# total predicted
r1_total <- r1[["predicted_totalBio"]]
r2_total <- r2[["predicted_totalBio"]]

# difference: model1 - model2
r_diff_total <- r1_total - r2_total
names(r_diff_total) <- "difference"

# per-group layers (cover-weighted)
weighted_lyrs <- names(r1)[!grepl("_potential|predicted_totalBio", names(r1))]
potential_lyrs <- names(r1)[grepl("_potential", names(r1))]
```

```{r sample-from-rasters}
# sample points from both rasters for scatterplots
set.seed(42)

r1a <- add_suffix2names(r1, '_m1')
r2a <- add_suffix2names(r2, '_m2')

r_stack <- c(r1a, r2a)

df_sample <- terra::spatSample(r_stack, size = params$n_sample_raster,
                                method = "random", na.rm = TRUE, xy = FALSE)

```

```{r print-config}
cat("Model type:", model_type, "\n\n")

cat("--- Model 1:", label1, "---\n")
cat("  Formula:", deparse(m1$config$model$formula), "\n")
cat("  Lambda:", signif(m1$cv$selected$lambda, 3), "\n")
cat("  N training:", m1$n_train, "\n\n")

cat("--- Model 2:", label2, "---\n")
cat("  Formula:", deparse(m2$config$model$formula), "\n")
cat("  Lambda:", signif(m2$cv$selected$lambda, 3), "\n")
cat("  N training:", m2$n_train, "\n")
```

## 1. CV Performance Comparison

```{r cv-comparison}
cv_summary <- function(m, label) {
  m$cv$selected |>
    dplyr::mutate(model = label) |>
    dplyr::select(model, lambda, mae_log1p, mae_log1p_se, rmse_log1p, mae, rmse, cor)
}

cv_compare <- dplyr::bind_rows(
  cv_summary(m1, label1),
  cv_summary(m2, label2)
) |>
  dplyr::mutate(dplyr::across(dplyr::where(is.numeric), \(x) signif(x, 4)))

knitr::kable(cv_compare, caption = "CV performance at selected lambda")
```


## 2. Total Biomass: Maps

### Side by Side

```{r maps-total-side-by-side, fig.width=14, fig.height=6}
# shared color limits
total_q99 <- max(
  terra::global(r1_total, fun = function(x) quantile(x, 0.99, na.rm = TRUE))[[1]],
  terra::global(r2_total, fun = function(x) quantile(x, 0.99, na.rm = TRUE))[[1]]
)
total_limits <- c(0, total_q99)

g1 <- plot_map_conus(r1_total,
                     colorscale = colorscale_biomass(limits = total_limits),
                     title = paste("Model 1:", label1))
g2 <- plot_map_conus(r2_total,
                     colorscale = colorscale_biomass(limits = total_limits),
                     title = paste("Model 2:", label2))
g1 + g2
```

### Difference (Model 1 − Model 2)

```{r map-total-diff, fig.width=10, fig.height=6}
diff_q99 <- terra::global(
  abs(r_diff_total),
  fun = function(x) quantile(x, 0.99, na.rm = TRUE)
)[[1]]

plot_map_conus(
  r_diff_total,
  colorscale = colorscale_diverging(
    name = "M1 − M2\n(Mg/ha)",
    limits = c(-diff_q99, diff_q99)
  ),
  title = paste("Difference in total biomass:", label1, "−", label2)
)
```

### Scatterplot: Model 1 vs Model 2

```{r scatter-total, fig.width=6, fig.height=5}
ggplot(df_sample, aes(x = predicted_totalBio_m1, y = predicted_totalBio_m2)) +
  geom_hex(bins = 100) +
  scale_fill_viridis_c(trans = "log10") +
  geom_abline(slope = 1, intercept = 0, color = "red", linetype = "dashed") +
  labs(x = paste("Model 1:", label1),
       y = paste("Model 2:", label2),
       title = "Total predicted biomass: Model 1 vs Model 2") +
  coord_equal()
```

## 3. Per-PFT Biomass: Cover-Weighted {.tabset}

```{r pft-weighted-comparison, fig.width=14, fig.height=10, eval=length(weighted_lyrs) > 0, results='asis'}
for (lyr in weighted_lyrs) {
  r1_lyr <- r1[[lyr]]
  r2_lyr <- r2[[lyr]]
  r_diff_lyr <- r1_lyr - r2_lyr
  
  lyr_q99 <- max(
    terra::global(r1_lyr, fun = function(x) quantile(x, 0.99, na.rm = TRUE))[[1]],
    terra::global(r2_lyr, fun = function(x) quantile(x, 0.99, na.rm = TRUE))[[1]]
  )
  
  diff_lyr_q99 <- terra::global(
    abs(r_diff_lyr),
    fun = function(x) quantile(x, 0.99, na.rm = TRUE)
  )[[1]]
  
  cat("\n###", lyr, "\n\n")
  
  # side by side maps
  g1 <- plot_map_conus(r1_lyr,
                       colorscale = colorscale_biomass(limits = c(0, lyr_q99)),
                       title = paste("M1:", label1))
  g2 <- plot_map_conus(r2_lyr,
                       colorscale = colorscale_biomass(limits = c(0, lyr_q99)),
                       title = paste("M2:", label2))
  g_diff <- plot_map_conus(r_diff_lyr,
                           colorscale = colorscale_diverging(
                             name = "M1−M2",
                             limits = c(-diff_lyr_q99, diff_lyr_q99)),
                           title = paste("Difference:", lyr))
  
  print(g1+g2)
  cat("\n\n")
  print(g_diff)
  cat("\n\n")
  
  # scatterplot
  m1_col <- paste0(lyr, "_m1")
  m2_col <- paste0(lyr, "_m2")
  if (all(c(m1_col, m2_col) %in% names(df_sample))) {
    g_scatter <- ggplot(df_sample, aes(x = .data[[m1_col]], y = .data[[m2_col]])) +
      geom_hex(bins = 80) +
      scale_fill_viridis_c(trans = "log10") +
      geom_abline(slope = 1, intercept = 0, color = "red", linetype = "dashed") +
      labs(x = paste("M1:", label1), y = paste("M2:", label2),
           title = paste(lyr, "— M1 vs M2")) +
      coord_equal()
    print(g_scatter)
    cat("\n\n")
  }
}
```

## 4. Per-PFT Biomass: Potential (Cover = 100%) {.tabset}

```{r pft-potential-comparison, fig.width=14, fig.height=10, eval=length(potential_lyrs) > 0, results='asis'}
for (lyr in potential_lyrs) {
  r1_lyr <- r1[[lyr]]
  r2_lyr <- r2[[lyr]]
  r_diff_lyr <- r1_lyr - r2_lyr
  
  lyr_q99 <- max(
    terra::global(r1_lyr, fun = function(x) quantile(x, 0.99, na.rm = TRUE))[[1]],
    terra::global(r2_lyr, fun = function(x) quantile(x, 0.99, na.rm = TRUE))[[1]]
  )
  
  diff_lyr_q99 <- terra::global(
    abs(r_diff_lyr),
    fun = function(x) quantile(x, 0.99, na.rm = TRUE)
  )[[1]]
  
  clean_name <- gsub("_potential$", " (potential)", lyr)
  
  cat("\n###", clean_name, "\n\n")
  
  g1 <- plot_map_conus(r1_lyr,
                       colorscale = colorscale_biomass(limits = c(0, lyr_q99),
                                                       option = "magma"),
                       title = paste("M1:", label1))
  g2 <- plot_map_conus(r2_lyr,
                       colorscale = colorscale_biomass(limits = c(0, lyr_q99),
                                                       option = "magma"),
                       title = paste("M2:", label2))
  g_diff <- plot_map_conus(r_diff_lyr,
                           colorscale = colorscale_diverging(
                             name = "M1−M2",
                             limits = c(-diff_lyr_q99, diff_lyr_q99)),
                           title = paste("Difference:", clean_name))
  
  print(g1+g2)
  cat("\n\n")
  print(g_diff)
  cat("\n\n")
  
  # scatterplot
  m1_col <- paste0(lyr, "_m1")
  m2_col <- paste0(lyr, "_m2")
  if (all(c(m1_col, m2_col) %in% names(df_sample))) {
    g_scatter <- ggplot(df_sample, aes(x = .data[[m1_col]], y = .data[[m2_col]])) +
      geom_hex(bins = 80) +
      scale_fill_viridis_c(trans = "log10") +
      geom_abline(slope = 1, intercept = 0, color = "red", linetype = "dashed") +
      labs(x = paste("M1:", label1), y = paste("M2:", label2),
           title = paste(clean_name, "— M1 vs M2")) +
      coord_equal()
    print(g_scatter)
    cat("\n\n")
  }
}
```

## 5. Coefficient Comparison

```{r coef-comparison}
cat("--- Model 1:", label1, "---\n")
cat("Alpha:\n")
print(round(m1$fit$par$alpha, 3))
cat("\nB:\n")
print(round(m1$fit$par$B, 3))
cat("Sigma:", round(m1$fit$par$sigma, 3), "\n\n")

cat("--- Model 2:", label2, "---\n")
cat("Alpha:\n")
print(round(m2$fit$par$alpha, 3))
cat("B:\n")
print(round(m2$fit$par$B, 3))
cat("\nSigma:", round(m2$fit$par$sigma, 3), "\n")
```

## 6. External Validation: BIGMAP AGB (Treed Areas)

Comparison of each model's tree cover-weighted biomass predictions against the
FIA BIGMAP AGB dataset, restricted to treed pixels.

```{r bigmap-setup, eval=is_woody}
# load CONUS rasters (needed for masks)
rasters <- load_conus_rasters()

r1_tree <- r1[["totalTree"]]
r2_tree <- r2[["totalTree"]]

bigmap1 <- compare_to_bigmap(r1_tree, rasters)
bigmap2 <- compare_to_bigmap(r2_tree, rasters)
```

### Metrics Comparison

```{r bigmap-metrics-comparison, eval=is_woody}
bigmap_compare <- bind_rows(
  bigmap1$metrics |> mutate(model = label1),
  bigmap2$metrics |> mutate(model = label2)
) |>
  mutate(value = signif(value, 4)) |>
  pivot_wider(names_from = model, values_from = value)

knitr::kable(bigmap_compare,
             caption = "Model vs BIGMAP AGB (treed areas)")
```

### Scatterplots

```{r bigmap-scatter-comparison, fig.width=12, fig.height=5, eval=is_woody}

g1_bigmap <- bigmap_scatter(bigmap1$sample) +
  labs(title = paste("Model 1:", label1)) 

g2_bigmap <- bigmap_scatter(bigmap1$sample) +
  labs(title = paste("Model 2:", label2)) 

g1_bigmap + g2_bigmap
```

## Session Info

```{r session-info}
cat("Model 1:", suffix1, "\n")
cat("Model 2:", suffix2, "\n")
cat("N sampled from rasters:", params$n_sample_raster, "\n")
cat("Rendered:", format(Sys.time(), "%Y-%m-%d %H:%M:%S"), "\n")
```
