---
title: "Spatial Diagnostics: Predicted Biomass Maps"
author: "Martin Holdrege"
date: "`r Sys.Date()`"
params:
  model_type: "woody"
  data_version: "d05"
  purer_version: "p04"
  model_version: "m09"
  cover_source: "rap"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
	echo = TRUE,
	message = FALSE,
	warning = FALSE
)
# dynamic figure dimensions based on model type
is_woody <- params$model_type == "woody"
fig_height_pft <- if (is_woody) 10 else 6
```

## Purpose

Wall-to-wall CONUS maps of observed and predicted biomass for the
`r params$model_type` model`r if (params$cover_source != "rap") paste0(" (cover source: ", params$cover_source, ")")`.
Includes per-PFT decomposition for woody models (cover-weighted and potential).

## Setup

```{r source-and-load}
source("Functions/init.R")
library(patchwork)
source_functions()
theme_set(theme_bw())

model_type <- params$model_type
vd <- params$data_version
vm <- params$model_version
vp <- params$purer_version
cover_source <- params$cover_source

cover_label <- if (cover_source != "rap") paste0(" [cover: ", cover_source, "]") else ""

suffix <- paste0(model_type, '_', vd, '-', vp, '-', vm)

# --- load fitted model ---
m <- load_fit(suffix)
config <- m$config
scale_df <- m$scale
cover_cols <- config$model$cover_cols
```

```{r load-rasters}
# --- load prediction raster ---
# created in Fit/02_predict_rasters.R
r_pred_all <- load_pred_raster(suffix, cover_source = cover_source)

cat("Prediction layers:", paste(names(r_pred_all), collapse = ", "), "\n")
cat("Cover source:", cover_source, "\n")

# --- load CONUS rasters (for observed + masks + cover) ---
rasters <- load_conus_rasters(cover_source = cover_source)

# observed biomass raster
r_observed <- if (model_type == "herb") {
  rasters$herb_bio
} else {
  rasters$gedi
}
names(r_observed) <- "observed"

# masks (same as used for training data)
mask_lcmap <- rasters$mask_lcmap
mask_fire <- rasters$mask_fire
```

```{r extract-layers}
# extract prediction layers
r_total_pred <- r_pred_all[["predicted_totalBio"]]

# per-group layers (cover-weighted)
weighted_lyrs <- names(r_pred_all)[!grepl("_potential|predicted_totalBio",
                                          names(r_pred_all))]
r_weighted <- if (length(weighted_lyrs) > 0) r_pred_all[[weighted_lyrs]] else NULL

# per-group layers (potential)
potential_lyrs <- names(r_pred_all)[grepl("_potential", names(r_pred_all))]
r_potential <- if (length(potential_lyrs) > 0) r_pred_all[[potential_lyrs]] else NULL

# --- mask observed for residual computation ---
r_obs_masked <- terra::mask(r_observed, mask_lcmap, maskvalues = 0)
if (model_type == "woody") {
  r_obs_masked <- terra::mask(r_obs_masked, mask_fire, maskvalues = 0)
}

# --- residual: observed - predicted (masked to valid area) ---
r_pred_masked <- terra::mask(r_total_pred, r_obs_masked)
r_diff <- r_pred_masked - r_obs_masked
names(r_diff) <- "residual"
```

```{r compute-limits}
# shared color limits for observed vs predicted comparison
obs_range <- terra::global(r_obs_masked, c("min", "max"), na.rm = TRUE)
pred_range <- terra::global(r_total_pred, c("min", "max"), na.rm = TRUE)
bio_limits <- c(
  min(obs_range$min, pred_range$min, na.rm = TRUE),
  max(obs_range$max, pred_range$max, na.rm = TRUE)
)
# cap upper limit at a reasonable percentile to avoid outlier stretching
bio_upper <- terra::global(
  c(r_obs_masked, r_total_pred), 
  fun = function(x) quantile(x, 0.99, na.rm = TRUE)
)
bio_limits[2] <- max(bio_upper[[1]])

# symmetric limits for residual
diff_range <- terra::global(r_diff, c("min", "max"), na.rm = TRUE)
diff_abs_max <- max(abs(diff_range$min), abs(diff_range$max), na.rm = TRUE)
diff_q99 <- terra::global(
  abs(r_diff), 
  fun = function(x) quantile(x, 0.99, na.rm = TRUE)
)
diff_limit <- min(diff_abs_max, diff_q99[[1]])
```

```{r print-config}
cat("Model type:", model_type, "\n")
cat("Model version:", vm, "\n")
cat("Data version:", vd, "\n")
cat("Formula:", deparse(config$model$formula), "\n")
cat("Cover cols:", paste(cover_cols, collapse = ", "), "\n")
cat("Biomass color limits:", round(bio_limits, 2), "\n")
cat("Residual limits: ±", round(diff_limit, 2), "\n")
```

## 1. Observed Biomass

```{r map-observed, fig.width=10, fig.height=6}
plot_map_conus(
  r_obs_masked,
  colorscale = colorscale_biomass(
    name = "Biomass\n(Mg/ha)",
    limits = bio_limits
  ),
  title = paste0("Observed ", model_type, " biomass")
)
```

## 1b. Cover Maps

```{r cover-maps, fig.width=10, fig.height=fig_height_pft}
cover_lyrs <- if (model_type == "woody") {
  c("totalTreeCov", "totalShrubCov")
} else {
  "totalHerbaceousCov"
}

cover_plots <- lapply(cover_lyrs, function(lyr) {
  r <- rasters$cover[[lyr]]
  limits <- c(0, raster_quantile(r, 0.99))
  plot_map_conus(
    r,
    colorscale = colorscale_biomass(name = "Cover\n(proportion)",
                                    limits = limits),
    title = lyr
  )
})

patchwork::wrap_plots(cover_plots, ncol = 1)
```

## 2. Predicted Total Biomass

```{r map-predicted, fig.width=10, fig.height=6}
plot_map_conus(
  r_total_pred,
  colorscale = colorscale_biomass(
    name = "Biomass\n(Mg/ha)",
    limits = bio_limits
  ),
  title = paste0("Predicted ", model_type, " biomass (wall-to-wall)")
)
```

## 3. Residual (Observed − Predicted)

Masked to areas where observed data exists (LCMAP valid pixels
`r if (model_type == "woody") "and unburned pixels"`).

```{r map-residual, fig.width=10, fig.height=6}
plot_map_conus(
  r_diff,
  colorscale = colorscale_diverging(
    name = "Pred - Obs \n(Mg/ha)",
    limits = c(-diff_limit, diff_limit)
  ),
  title = paste0("Residual: observed − predicted ", model_type, " biomass")
)
```

## 4. Per-PFT Biomass (Cover-Weighted)

```{r pft-weighted-maps, fig.width=10, fig.height=fig_height_pft, eval=!is.null(r_weighted)}
# individual maps per PFT with independent color scales
pft_weighted_plots <- lapply(weighted_lyrs, function(lyr) {
  r <- r_pred_all[[lyr]]
  lyr_max <- raster_quantile(r, 0.99) 
  
  plot_map_conus(
    r,
    colorscale = colorscale_biomass(
      name = "Biomass\n(Mg/ha)",
      limits = c(0, lyr_max[[1]])
    ),
    title = lyr
  )
})

patchwork::wrap_plots(pft_weighted_plots, ncol = 1)
```

## 5. Per-PFT Potential Biomass (Cover = 100%)

```{r pft-potential-maps, fig.width=10, fig.height=fig_height_pft, eval=!is.null(r_potential)}
pft_potential_plots <- lapply(potential_lyrs, function(lyr) {
  r <- r_pred_all[[lyr]]
  lyr_max <- terra::global(r, fun = function(x) quantile(x, 0.99, na.rm = TRUE))
  
  plot_map_conus(
    r,
    colorscale = colorscale_biomass(
      name = "Potential\n(Mg/ha)",
      limits = c(0, lyr_max[[1]]),
      option = "magma"
    ),
    title = gsub("_potential$", " (potential, cover=100%)", lyr)
  )
})

patchwork::wrap_plots(pft_potential_plots, ncol = 1)
```

## 6. Sagebrush Biome Focus

```{r load-sagebrush-mask}
r_sage_mask <- load_sagebrush_mask()

sage_ext <- terra::ext(r_sage_mask)
r_sage_mask <- terra::crop(r_sage_mask, sage_ext)
```

```{r mask-to-sagebrush}

crop_mask_sage <- function(r) {
  terra::crop(r, sage_ext) |>
    terra::mask(r_sage_mask)
}
# mask and crop predictions to sagebrush region
r_total_sage <- crop_mask_sage(r_total_pred)

# per-group weighted
r_weighted_sage <- crop_mask_sage(r_pred_all[[weighted_lyrs]])


# per-group potential
r_potential_sage <- crop_mask_sage(r_pred_all[[potential_lyrs]])

# observed
r_obs_sage <- crop_mask_sage(r_obs_masked)
```


### Observed Biomass

```{r sage-observed, fig.width=8, fig.height=6}

q <- 0.9
max_sage <- max(c(raster_quantile(r_obs_sage, q),
                      raster_quantile(r_total_sage, q)))

cap1 <- paste0('colors trimmed to ', q*100, 'th percentile')
plot_map_conus(r_obs_sage,
               colorscale = colorscale_biomass(limits = c(0, max_sage)),
               title = paste("Observed", model_type, "biomass — Sagebrush biome")) +
  labs(caption = cap1)
```

### Total Predicted Biomass

```{r sage-total-pred, fig.width=8, fig.height=6}
plot_map_conus(r_total_sage,
               colorscale = colorscale_biomass(limits = c(0, max_sage)),
               title = paste("Predicted", model_type, "biomass — Sagebrush biome"))+
  labs(caption = cap1)
```

### Per-PFT Biomass (Cover-Weighted) {.tabset}

```{r sage-pft-weighted, fig.width=8, fig.height=6, results='asis'}
q2 <- 0.95
cap2 <- paste0('color trimmed to ', q2*100, 'th percentile')
for (lyr in weighted_lyrs) {
  
  r_lyr <- r_weighted_sage[[lyr]]
  lyr_q <- raster_quantile(r_lyr, q2)
  
  
  cat("\n####", lyr, "\n\n")
  
  g <- plot_map_conus(r_lyr,
                      colorscale = colorscale_biomass(limits = c(0, lyr_q)),
                      title = paste(lyr, "— Sagebrush biome")) +
    labs(caption = cap2)
  print(g)
  cat("\n\n")
  
  # histogram
  vals <- terra::values(r_lyr, na.rm = TRUE)
  g_hist <- ggplot(data.frame(biomass = vals[,1]), aes(x = biomass)) +
    geom_histogram(bins = 80, fill = "steelblue", color = "white") +
    labs(x = paste(lyr, "(Mg/ha)"), y = "Count",
         title = paste(lyr, "— distribution in sagebrush biome")) +
    scale_x_continuous(trans = "log1p")
  print(g_hist)
  cat("\n\n")
}
```

### Per-PFT Biomass (Potential) {.tabset}

```{r sage-pft-potential, fig.width=8, fig.height=6, results='asis'}
for (lyr in potential_lyrs) {
  r_lyr <- r_potential_sage[[lyr]]
  lyr_q <- raster_quantile(r_lyr, q2)
  
  clean_name <- gsub("_potential$", " (potential)", lyr)
  cat("\n####", clean_name, "\n\n")
  
  g <- plot_map_conus(r_lyr,
                      colorscale = colorscale_biomass(limits = c(0, lyr_q),
                                                      option = "magma"),
                      title = paste(clean_name, "— Sagebrush biome")) +
    labs(caption = cap2)
  print(g)
  cat("\n\n")
  
  vals <- terra::values(r_lyr, na.rm = TRUE)
  g_hist <- ggplot(data.frame(biomass = vals[,1]), aes(x = biomass)) +
    geom_histogram(bins = 80, fill = "steelblue", color = "white") +
    labs(x = paste(clean_name, "(Mg/ha)"), y = "Count",
         title = paste(clean_name, "— distribution in sagebrush biome"))
  print(g_hist)
  cat("\n\n")
}
```



## 7. Comparison with BIGMAP AGB (Treed Areas)

External validation of woody model predictions against the FIA BIGMAP
aboveground biomass dataset (2018), restricted to pixels classified as treed.
Uses the tree cover-weighted prediction layer (`totalTree`) rather than total
biomass, since BIGMAP is tree AGB and total includes shrub biomass.

```{r bigmap-load, eval=is_woody}
r_tree_pred <- r_pred_all[["totalTree"]]
bigmap <- compare_to_bigmap(r_tree_pred, rasters)

# shared color limits for maps
bigmap_q99 <- max(
  terra::global(bigmap$r_bigmap_masked, fun = function(x) quantile(x, 0.99, na.rm = TRUE))[[1]],
  terra::global(bigmap$r_pred_masked, fun = function(x) quantile(x, 0.99, na.rm = TRUE))[[1]]
)
bigmap_limits <- c(0, bigmap_q99)
```

### Side-by-Side Maps

```{r bigmap-maps, fig.width=14, fig.height=6, eval=is_woody}
g_bigmap <- plot_map_conus(
  bigmap$r_bigmap_masked,
  colorscale = colorscale_biomass(name = "Biomass\n(Mg/ha)",
                                  limits = bigmap_limits),
  title = "BIGMAP AGB (treed areas)"
)
g_pred_treed <- plot_map_conus(
  bigmap$r_pred_masked,
  colorscale = colorscale_biomass(name = "Biomass\n(Mg/ha)",
                                  limits = bigmap_limits),
  title = "Model predicted tree biomass (treed areas)"
)
g_bigmap + g_pred_treed
```

### Difference Map (Model Predicted − BIGMAP)

```{r bigmap-diff-map, fig.width=10, fig.height=6, eval=is_woody}
r_bigmap_diff <- bigmap$r_pred_masked - bigmap$r_bigmap_masked
names(r_bigmap_diff) <- "difference"

bigmap_diff_q99 <- terra::global(
  abs(r_bigmap_diff),
  fun = function(x) quantile(x, 0.99, na.rm = TRUE)
)[[1]]

plot_map_conus(
  r_bigmap_diff,
  colorscale = colorscale_diverging(
    name = "Pred − BIGMAP\n(Mg/ha)",
    limits = c(-bigmap_diff_q99, bigmap_diff_q99)
  ),
  title = "Difference: BIGMAP AGB (treed areas) - model predicted"
)
```

### Scatterplot

```{r bigmap-scatter, fig.width=6, fig.height=5, eval=is_woody}
ggplot(bigmap$sample, aes(x = bigmap, y = predicted)) +
  geom_hex(bins = 100) +
  scale_fill_viridis_c(trans = "log10") +
  geom_abline(slope = 1, intercept = 0, color = "red", linetype = "dashed") +
  labs(x = "BIGMAP AGB (Mg/ha)", y = "Model predicted tree biomass (Mg/ha)",
       title = "Model vs BIGMAP AGB (treed areas)") +
  coord_equal()
```

### Summary Metrics

```{r bigmap-metrics, eval=is_woody}
bigmap$metrics |>
  mutate(value = signif(value, 4)) |>
  knitr::kable(caption = "Model vs BIGMAP AGB summary (treed areas)")
```

## Session Info

```{r session-info}
cat("Suffix:", suffix, "\n")
cat("Cover source:", cover_source, "\n")
cat("Rendered:", format(Sys.time(), "%Y-%m-%d %H:%M:%S"), "\n")
```
