---
title: "Climate and soil predictors: PCA and correlations"
author: "Martin Holdrege"
date: "`r Sys.Date()`"
params:
  vc: null        # cover data version; null uses opt$vc
  n_top: 8        # top variables per PC to report
  n_select: 15    # variables to add in forward selection
  n_conus: 100000 # CONUS cells to sample for section 1
  test_run: FALSE
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

## Purpose

Help choose a small set of predictors that covers most of the climate and soil
variation, is biologically reasonable, and isn't strongly correlated. Help
decide which variables will be selected from when model fitting. Done
twice:

1. Normals and soils, on a random sample of available CONUS cells (LCMAP and
   2023 fire masks), so one selection serves both the cover and biomass
   models. The chosen set is also checked on the cover training data.
2. Normals, soils and 3-yr anomalies, on the cover training data (anomalies
   exist only there; cover models only).

**Coverage** of a set = RM² from `subselect::rm.coef()`: the share of the
total variance of all variables explained by regressing them on the set
(McCabe's criterion for "principal variables"). Comparable to the cumulative
variance of the same number of PCs, which is the best possible. **Best
subsets** come from `subselect::improve()` with criterion "RM": for each size
k, the k variables with the highest coverage (a local search, so not
guaranteed optimal). See `?subselect::rm.coef` and `?subselect::improve`.

## Setup

```{r setup-data}
source("Functions/init.R")
source_functions()  # read_cover_training(), variable_selection.R
theme_set(theme_classic())

vc <- if (is.null(params$vc)) opt$vc else params$vc

dat <- read_cover_training(vc, normalize = TRUE)
if (params$test_run) dat <- slice_sample(dat, n = 5000)

soil_vars <- c("soilDepth", "clay_surface", "clay", "sand", "coarse",
               "carbon", "awc")
anom_vars <- str_subset(names(dat), "_3yrAnom$")
clim_vars <- c(str_remove(anom_vars, "_3yrAnom$"), soil_vars)
all_vars  <- c(clim_vars, anom_vars)

dat <- dat[complete.cases(dat[all_vars]), ]

# available CONUS cells: normals and soils only
n_conus <- if (params$test_run) 2000 else params$n_conus
set.seed(1)
conus <- read_climate_raster("current") |>
  align_raster(read_mask()) |>
  terra::mask(read_available_mask(2023)) |>
  terra::spatSample(size = n_conus, method = "random", na.rm = TRUE) |>
  standardize(vars = clim_vars, scale_df = read_scale_params())
conus <- conus$data[complete.cases(conus$data[clim_vars]), ]
```

CONUS sample: `r nrow(conus)` available cells. Cover data `r vc`:
`r nrow(dat)` complete pixel-years (repeat years at a pixel kept) at
`r n_distinct(dat$cell)` pixels.

```{r helpers}
plot_scree <- function(pca) {
  ggplot(pca_variance(pca), aes(pc_num)) +
    geom_col(aes(y = prop_var), fill = "steelblue", alpha = 0.7) +
    geom_line(aes(y = cum_var)) +
    geom_point(aes(y = cum_var)) +
    geom_hline(yintercept = 0.9, linetype = "dashed", color = "grey50") +
    scale_y_continuous(labels = scales::percent) +
    labs(x = "Principal component", y = "Variance (bars: per PC, line: cumulative)")
}

# PCs up to 95% cumulative variance
plot_loadings <- function(pca) {
  n_pc <- min(which(pca_variance(pca)$cum_var >= 0.95))
  ggplot(pca_loadings_long(pca, n_pc), aes(PC, variable, fill = loading)) +
    geom_tile(color = "white") +
    geom_text(aes(label = sprintf("%.2f", loading)), size = 2.2) +
    scale_fill_gradient2(low = "#b2182b", high = "#2166ac", limits = c(-1, 1)) +
    labs(x = NULL, y = NULL)
}

top_loadings <- function(pca) {
  n_pc <- min(which(pca_variance(pca)$cum_var >= 0.95))
  pca_loadings_long(pca, n_pc) |>
    group_by(PC) |>
    slice_max(abs(loading), n = params$n_top, with_ties = FALSE) |>
    mutate(rank = row_number(),
           entry = paste0(variable, " (", sprintf("%+.2f", loading), ")")) |>
    select(rank, PC, entry) |>
    pivot_wider(names_from = PC, values_from = entry)
}

plot_cor <- function(cor_mat) {
  ggplot(cor_long(cor_mat), aes(var1, var2, fill = r)) +
    geom_tile(color = "white") +
    scale_fill_gradient2(low = "#b2182b", high = "#2166ac", limits = c(-1, 1)) +
    labs(x = NULL, y = NULL) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
}

kable_set <- function(tbl, caption) {
  tbl |>
    mutate(cum_coverage = scales::percent(cum_coverage, accuracy = 0.1)) |>
    knitr::kable(digits = 2, caption = caption)
}
```

## 1. Normals and soils (CONUS sample)

```{r clim-pca}
pca_clim <- prcomp(conus[clim_vars], scale. = TRUE)
cor_clim <- cor(conus[clim_vars])
```

```{r clim-exclude}
# improve() needs a non-singular correlation matrix, so variables that are
# (near-)exact linear combinations of others are left out of the subset
# search. MAT is presumably the mean of T_min and T_max, so one of the three
# goes (edit to choose).
exclude_clim <- "T_min"

# any remaining dependency; trim.matrix() names the variable(s) to add to
# exclude_clim (see ?subselect::trim.matrix)
keep <- setdiff(clim_vars, exclude_clim)
remaining <- subselect::trim.matrix(cor_clim[keep, keep],
                                    tolval = 1000 * .Machine$double.eps)$names.discarded
stopifnot(length(remaining) == 0)
```

### PCA

```{r clim-scree, fig.width = 7, fig.height = 4}
plot_scree(pca_clim)
```

```{r clim-loadings, fig.width = 8, fig.height = 8}
plot_loadings(pca_clim)
knitr::kable(top_loadings(pca_clim),
             caption = paste("Top", params$n_top, "variables per PC by |loading| (showing PCs that explain 95%)"))
```

### Correlations

Variables clustered on 1 - |r|.

```{r clim-cor, fig.width = 9, fig.height = 8}
plot_cor(cor_clim)
```

### Candidate sets

Best subsets by coverage (a data-driven benchmark, not a recommendation),
with the cumulative variance of the same number of PCs for comparison. For
your candidate set, `cum_coverage` is the coverage as each variable is added,
`max_abs_r` the largest |r| with the rest of the set.

```{r clim-sets}
best_subsets(cor_clim, params$n_select, exclude = exclude_clim) |>
  mutate(pca = pca_variance(pca_clim)$cum_var[k],
         across(c(coverage, pca), \(x) scales::percent(x, accuracy = 0.1))) |>
  knitr::kable(caption = "Best subsets (subselect::improve, RM)")

# edit to taste (starting point: Alice's CONUS set)
candidate_clim <- c("MAT", "MAP", "PrecipTempCorr", "isothermality",
                    "WDD_mean", "clay_surface", "coarse", "awc")
summarise_set(conus[clim_vars], candidate_clim) |>
  kable_set("Candidate set")

# same set, in the cover training data
summarise_set(dat[clim_vars], candidate_clim) |>
  kable_set("Candidate set, cover training data")
```

```{r clim-subsets}
# each item: k = one or more subset sizes; include = NULL or variables that
# must be in every subset
subset_specs <- list(
  list(k = 4:8, include = NULL),
  list(k = 5:7,   include = c("MAT", "MAP", "PrecipTempCorr", 'awc')),
  list(k = 6:8,   include = c("MAT", "MAP", "PrecipTempCorr", 'isothermality', 
                              'awc'))
)

subset_results <- map(subset_specs, \(s) {
  stopifnot(all(s$k > length(s$include)))  # k must exceed the required count
  best_subsets(cor_clim, kmax = max(s$k), include = s$include,
               exclude = exclude_clim) |>
    filter(k %in% s$k) |>
    mutate(required = if (is.null(s$include)) "none" else paste(s$include, collapse = ", "),
           .before = 1)
}) |>
  bind_rows()

subset_results |>
  mutate(pca = pca_variance(pca_clim)$cum_var[k],
         across(c(coverage, pca), \(x) scales::percent(x, accuracy = 0.1))) |>
  knitr::kable(caption = "Best subsets for given sizes and required variables")
```

## 2. Normals, soils and anomalies

```{r both-pca}
pca_both <- prcomp(dat[all_vars], scale. = TRUE)
cor_both <- cor(dat[all_vars])
```

```{r both-exclude}
# as in section 1; the same dependency presumably holds for the anomalies
exclude_both <- c(exclude_clim, "T_min_3yrAnom")

keep <- setdiff(all_vars, exclude_both)
remaining <- subselect::trim.matrix(cor_both[keep, keep],
                                    tolval = 1000 * .Machine$double.eps)$names.discarded
stopifnot(length(remaining) == 0)
```

### Anomalies vs normals and soils

```{r both-cross, fig.width = 9, fig.height = 7}
as_tibble(cor_both[anom_vars, clim_vars], rownames = "anomaly") |>
  pivot_longer(-anomaly, names_to = "variable", values_to = "r") |>
  ggplot(aes(variable, anomaly, fill = r)) +
  geom_tile(color = "white") +
  scale_fill_gradient2(low = "#b2182b", high = "#2166ac", limits = c(-1, 1)) +
  labs(x = NULL, y = NULL) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

### PCA

```{r both-scree, fig.width = 7, fig.height = 4}
plot_scree(pca_both)
```

```{r both-loadings, fig.width = 9, fig.height = 11}
plot_loadings(pca_both)
knitr::kable(top_loadings(pca_both),
             caption = paste("Top", params$n_top, "variables per PC by |loading|"))
```

### Correlations among anomalies

```{r both-cor, fig.width = 9, fig.height = 8}
plot_cor(cor_both[anom_vars, anom_vars])
```

### Candidate sets

```{r both-sets}
best_subsets(cor_both, params$n_select, exclude = exclude_both) |>
  mutate(pca = pca_variance(pca_both)$cum_var[k],
         across(c(coverage, pca), \(x) scales::percent(x, accuracy = 0.1))) |>
  knitr::kable(caption = "Best subsets (subselect::improve, RM)")

# edit to taste
candidate_both <- c(candidate_clim, "MAT_3yrAnom", "MAP_3yrAnom")
summarise_set(dat[all_vars], candidate_both) |>
  kable_set("Candidate set")
```

```{r both-subsets}
# each item: k = one or more subset sizes; include = NULL or variables that
# must be in every subset
subset_specs <- list(
  list(k = 6:12, include = NULL),
  list(k = 6:12,   include = c("MAT", "MAP", "PrecipTempCorr", 'awc'))
)

subset_results <- map(subset_specs, \(s) {
  stopifnot(all(s$k > length(s$include)))  # k must exceed the required count
  best_subsets(cor_both, kmax = max(s$k), include = s$include,
               exclude = exclude_both) |>
    filter(k %in% s$k) |>
    mutate(required = if (is.null(s$include)) "none" else paste(s$include, collapse = ", "),
           .before = 1)
}) |>
  bind_rows()

subset_results |>
  mutate(pca = pca_variance(pca_both)$cum_var[k],
         across(c(coverage, pca), \(x) scales::percent(x, accuracy = 0.1))) |>
  knitr::kable(caption = "Best subsets for given sizes and required variables")
```

## Session info

```{r session-info}
sessionInfo()
```
