---
title: "Climate and soil predictors: PCA and correlations"
author: "Martin Holdrege"
date: "`r Sys.Date()`"
params:
  vc: null        # cover data version; null uses opt$vc
  n_top: 8        # top variables per PC to report
  max_k: 15       # largest subset size in the first best-subsets spec
  n_conus: 100000 # CONUS cells to sample for section 1
  test_run: FALSE
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

## Purpose

Help choose a small set of predictors that covers most of the climate and soil
variation, is biologically reasonable, and isn't strongly correlated. Help
decide which variables will be selected from when model fitting. Done
twice:

1. Normals and soils, on a random sample of available CONUS cells (LCMAP and
   2023 fire masks), so one selection serves both the cover and biomass
   models. The chosen set is also checked on the cover training data.
2. Normals, soils and 3-yr anomalies, on the cover training data (anomalies
   exist only there; cover models only).

**Coverage** of a set = RM² from `subselect::rm.coef()`: the share of the
total variance of all variables explained by regressing them on the set
(McCabe's criterion for "principal variables"). It's on the same scale as the
cumulative variance of the same number of PCs, which is the best possible.
**Best subsets** come from `subselect::improve()` with criterion "RM": for
each size k, the k variables with the highest coverage found by a local
search (not guaranteed optimal). See `?subselect::rm.coef` and
`?subselect::improve`. The functions used here are in
`variable_selection.R`.

## Setup

```{r setup-data}
source("Functions/init.R")
source_functions()  # read_cover_training(), variable_selection.R
theme_set(theme_classic())

vc <- if (is.null(params$vc)) opt$vc else params$vc

# PCA (with scale. = TRUE) and correlations don't depend on normalization, so
# data are used in their original units

# cover training data
dat <- read_cover_training(vc)
if (params$test_run) dat <- slice_sample(dat, n = 5000)

# variable groups
soil_vars <- c("soilDepth", "clay_surface", "clay", "sand", "coarse",
               "carbon", "awc")
anom_vars <- str_subset(names(dat), "_3yrAnom$")
clim_vars <- c(str_remove(anom_vars, "_3yrAnom$"), soil_vars)  # normals + soils
all_vars  <- c(clim_vars, anom_vars)

dat <- dat[complete.cases(dat[all_vars]), ]

# random sample of available CONUS cells (normals and soils only)
n_conus <- if (params$test_run) 2000 else params$n_conus

r_clim <- read_climate_raster("current") |>  # normals + soils, short names
  align_raster(read_mask())
r_available <- terra::mask(r_clim, read_available_mask(2023))

set.seed(1)
conus <- terra::spatSample(r_available, size = n_conus, method = "random",
                           na.rm = TRUE)
conus <- conus[complete.cases(conus[clim_vars]), clim_vars]
```

CONUS sample: `r nrow(conus)` available cells. Cover data `r vc`:
`r nrow(dat)` complete pixel-years (repeat years at a pixel kept) at
`r n_distinct(dat$cell)` pixels.

```{r helpers}
plot_scree <- function(pca) {
  ggplot(pca_variance(pca), aes(pc_num)) +
    geom_col(aes(y = prop_var), fill = "steelblue", alpha = 0.7) +
    geom_line(aes(y = cum_var)) +
    geom_point(aes(y = cum_var)) +
    geom_hline(yintercept = 0.9, linetype = "dashed", color = "grey50") +
    scale_y_continuous(labels = scales::percent) +
    labs(x = "Principal component", y = "Variance (bars: per PC, line: cumulative)")
}

plot_loadings <- function(pca, n_pc) {
  ggplot(pca_loadings_long(pca, n_pc), aes(PC, variable, fill = loading)) +
    geom_tile(color = "white") +
    geom_text(aes(label = sprintf("%.2f", loading)), size = 2.2) +
    scale_fill_gradient2(low = "#b2182b", high = "#2166ac", limits = c(-1, 1)) +
    labs(x = NULL, y = NULL)
}

plot_cor <- function(cor_mat) {
  ggplot(cor_long(cor_mat), aes(var1, var2, fill = r)) +
    geom_tile(color = "white") +
    scale_fill_gradient2(low = "#b2182b", high = "#2166ac", limits = c(-1, 1)) +
    labs(x = NULL, y = NULL) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
}

# best subsets, with the cumulative variance of the same number of PCs
kable_subsets <- function(subsets, pca, caption) {
  subsets |>
    mutate(pca = pca_variance(pca)$cum_var[k],
           coverage = scales::percent(coverage, accuracy = 0.1),
           pca = scales::percent(pca, accuracy = 0.1)) |>
    knitr::kable(caption = caption)
}

kable_set <- function(set_summary, caption) {
  set_summary |>
    mutate(cum_coverage = scales::percent(cum_coverage, accuracy = 0.1)) |>
    knitr::kable(digits = 2, caption = caption)
}
```

## 1. Normals and soils (CONUS sample)

```{r clim-pca}
pca_clim <- prcomp(conus[clim_vars], scale. = TRUE)
cor_clim <- cor(conus[clim_vars])
n_pc_clim <- n_pcs_needed(pca_clim, prop = 0.95)  # PCs shown below
```

### PCA

```{r clim-scree, fig.width = 7, fig.height = 4}
plot_scree(pca_clim)
```

```{r clim-loadings, fig.width = 8, fig.height = 8}
plot_loadings(pca_clim, n_pc_clim)

top_loadings(pca_clim, n_pc = n_pc_clim, n_top = params$n_top) |>
  knitr::kable(caption = paste("Top", params$n_top,
                               "variables per PC by |loading| (PCs that explain 95%)"))
```

### Correlations

Variables clustered on 1 - |r|.

```{r clim-cor, fig.width = 9, fig.height = 8}
plot_cor(cor_clim)
```

### Best subsets

A data-driven benchmark, not a recommendation. `pca` is the cumulative
variance of the same number of PCs.

```{r clim-subsets}
# MAT is the mean of T_min and T_max, so one of the three is left out of the
# search (edit to choose). best_subsets() stops and names any other
# (near-)exact linear dependency.
exclude_clim <- "T_min"

# each spec: subset sizes, and NULL or variables required in every subset
specs_clim <- list(
  list(sizes = 1:params$max_k, include = NULL),
  list(sizes = 5:params$max_k, 
     include = c("MAT", "MAP", "PrecipTempCorr", "awc"),
     exclude_improve = NULL),
  list(sizes = 5:params$max_k, 
       include = c("MAT", "MAP", "PrecipTempCorr", "awc"),
       exclude_improve = c('soilDepth', 'sand', 'clay')), 
  list(sizes = 5:params$max_k, 
     include = c("MAT", "MAP", "PrecipTempCorr", "awc"),
     # allow clay_surface, but not the other soil variables, not bias
     # toward climate variables
     exclude_improve = c('soilDepth', 'sand', 'clay', 'coarse', 'carbon')) 
)

best_subsets_by_spec(cor_clim, specs_clim, exclude = exclude_clim) |>
  kable_subsets(pca_clim, caption = "Best subsets")
```

### Candidate set

For each variable, in the order listed: `cum_coverage` is the coverage of the
set up to that variable, `max_abs_r` its largest |r| with the rest of the set,
`vif` its VIF within the set.

```{r clim-candidate}
# edit to taste (starting point)
candidate_clim <- c("MAT", "MAP", "PrecipTempCorr", "isothermality",
                                "awc", "carbon", 'WD_mean')

summarise_set(conus[clim_vars], candidate_clim) |>
  kable_set("Candidate set, CONUS sample")

summarise_set(dat[clim_vars], candidate_clim) |>
  kable_set("Candidate set, cover training data")
```

## 2. Normals, soils and anomalies (cover training data)

```{r both-pca}
pca_both <- prcomp(dat[all_vars], scale. = TRUE)
cor_both <- cor(dat[all_vars])
n_pc_both <- n_pcs_needed(pca_both, prop = 0.95)
```

### Anomalies vs normals and soils

```{r both-cross, fig.width = 9, fig.height = 7}
as_tibble(cor_both[anom_vars, clim_vars], rownames = "anomaly") |>
  pivot_longer(-anomaly, names_to = "variable", values_to = "r") |>
  ggplot(aes(variable, anomaly, fill = r)) +
  geom_tile(color = "white") +
  scale_fill_gradient2(low = "#b2182b", high = "#2166ac", limits = c(-1, 1)) +
  labs(x = NULL, y = NULL) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

### PCA

```{r both-scree, fig.width = 7, fig.height = 4}
plot_scree(pca_both)
```

```{r both-loadings, fig.width = 9, fig.height = 11}
plot_loadings(pca_both, n_pc_both)

top_loadings(pca_both, n_pc = n_pc_both, n_top = params$n_top) |>
  knitr::kable(caption = paste("Top", params$n_top,
                               "variables per PC by |loading| (PCs that explain 95%)"))
```

### Correlations among anomalies

```{r both-cor, fig.width = 9, fig.height = 8}
plot_cor(cor_both[anom_vars, anom_vars])
```

### Best subsets

```{r both-subsets}
# as in section 1; the same dependency presumably holds for the anomalies
exclude_both <- c(exclude_clim, "T_min_3yrAnom")

specs_both <- list(
  list(sizes = 6:12, include = NULL),
  list(sizes = 6:12, include = c("MAT", "MAP", "PrecipTempCorr", "awc"))
)

best_subsets_by_spec(cor_both, specs_both, exclude = exclude_both) |>
  kable_subsets(pca_both, caption = "Best subsets")
```

### Candidate set with paired anomalies

An anomaly is only used alongside its normal, so this adds the anomaly of
each normal in the section 1 candidate set. `max_abs_r` and `vif` for the
anomalies show whether they can be added without strong collinearity.

```{r both-candidate}
# anomalies of the chosen normals (soils have none)
paired_anoms <- intersect(paste0(candidate_clim, "_3yrAnom"), anom_vars)
candidate_both <- c(candidate_clim, paired_anoms)

summarise_set(dat[all_vars], candidate_both) |>
  kable_set("Candidate set with paired anomalies")
```

## Session info

```{r session-info}
sessionInfo()
```