---
title: "cwexp Model Pipeline: Proof of Concept"
author: "Martin Holdrege"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

## Purpose

This document walks through the full cwexp (what I'm calling 'cover weighted exponential' model) 
modeling pipeline on simulated data
where we know the truth. The goals are to:

1. Confirm the pipeline machinery works end-to-end.
2. Show that regularization helps when data are noisy and predictors include
   collinear junk variables.
3. Demonstrate parameter recovery and comparison to truth.
4. Serve as a reference for the function call sequences.

## Setup

```{r source-functions, warning = FALSE}
# source project code
source('Functions/init.R')
source_functions()
library(ggplot2)
library(dplyr)
theme_set(theme_bw())
```

## 1. Simulate Data


```{r simulate}
set.seed(42)
n <- 2000
dummy <- cwexp_make_dummy_data(n = n*2, sigma = 0.4, n_junk_pred = 3)
dat <- dummy$data[1:n, ]
 # not a traditional test set, but just for seeing how well model's recover the truth
dat_test <- dummy$data[(n+1):(2*n), ]
cover_cols <- dummy$spec$cover_cols
formula_full <- dummy$spec$formula  # totalBio ~ tmean + ppt + vpd + sand + junk1 + junk2 + junk3
pred_vars <- dummy$prep$x_cols

cat("Formula:", deparse(formula_full), "\n")
cat("Cover columns:", paste(cover_cols, collapse = ", "), "\n")
cat("True sigma:", dummy$par$sigma, "\n")
cat("Dimensions of B:", paste(dim(dummy$par$B), collapse = " x "), "\n")
```

True B matrix (rows = predictors, columns = cover groups). 

```{r true-B}
round(dummy$par$B, 3)
```

## 2. Compile TMB Model

```{r compile-tmb}
dll_en <- cwexp_tmb_compile("src/cwexp_lognormal_en_tmb.cpp", quiet = TRUE)
```

## 3. Single Unpenalized Fit

First, fit with `lambda = 0` (no penalty) to see baseline recovery.

```{r fit-unpenalized, message = FALSE}

fit0 <- cwexp_fit_tmb(
  data = dat,
  formula = formula_full,
  cover_cols = cover_cols,
  dll = dll_en,
  penalty = "elastic_net",
  en_alpha = 0.5,
  lambda = 0
)

# predicting on held out data
mu_true <- predict(dummy, newdata = dat_test, type = "mu")
mu_hat0 <- predict(fit0, newdata = dat_test, type = "mu")
```

Unpenalized fit


```{r}

cat("  cor(true mu, fitted mu):", round(cor(mu_true, mu_hat0), 3), "\n")
cat("  Estimated sigma:", round(fit0$par$sigma, 3),
    " (true:", dummy$par$sigma, ")\n")
```

```{r plot-unpenalized, fig.width=5, fig.height=4.5}

plot(mu_true, dat_test$totalBio,
     ylab = 'simulated biomass', xlab = 'true mu')
plot(log(mu_true), log(dat_test$totalBio),
     ylab = 'log(simulated biomass)', xlab = 'log(true mu))')
plot(mu_true, mu_hat0, pch = 16, cex = 0.4, col = rgb(0, 0, 0, 0.3),
     xlab = "True mu", ylab = "Fitted mu",
     main = "Unpenalized fit (lambda = 0)")
abline(0, 1, col = "red", lwd = 2)
```

Compare estimated B to true B:

```{r compare-B-unpenalized, fig.width=6, fig.height=5}
B_true_vec <- as.numeric(dummy$par$B)
B_hat0_vec <- as.numeric(fit0$par$B)
n_real <- length(c("tmean", "ppt", "vpd", "sand"))
n_total <- nrow(dummy$par$B)
is_junk <- rep(c(rep(FALSE, n_real), rep(TRUE, n_total - n_real)),
               ncol(dummy$par$B))

plot(B_true_vec, B_hat0_vec, pch = 16,
     col = ifelse(is_junk, "red", "black"),
     xlab = "True B", ylab = "Estimated B",
     main = "B recovery (unpenalized)")
abline(0, 1, col = "grey50", lty = 2)
abline(h = 0, v = 0, col = "grey80")
legend("topleft", legend = c("Real predictors", "Junk predictors"),
       col = c("black", "red"), pch = 16, bty = "n")
```

## 4. Lambda Path

Fit models across a range of lambda values to visualize the regularization 
path.

```{r lambda-path}

lam_max <- cwexp_lambda_max_l1_tmb(
  data = dat,
  formula = formula_full,
  cover_cols = cover_cols,
  dll_en = dll_en
)

en_alpha <- 0.5

lam_seq <- cwexp_lambda_path(
  lambda_max_l1 = lam_max$lambda_max_l1,
  en_alpha = en_alpha,
  n_lambda = 15,
  include_zero = TRUE
)

path_fit <- cwexp_fit_lambda_path_tmb(
  data = dat,
  formula = formula_full,
  cover_cols = cover_cols,
  dll = dll_en,
  en_alpha = en_alpha,
  lambda_seq = lam_seq
)

```

### Coefficient paths

```{r coef-paths, fig.width=8, fig.height=5}
df_B <- cwexp_extract_B_path(path_fit)
# tag junk vs real
df_B$is_junk <- grepl("^junk", df_B$predictor)

ggplot(df_B, aes(x = lambda, y = B, group = interaction(predictor, group))) +
  geom_line(aes(color = is_junk), alpha = 0.6) +
  scale_color_manual(values = c("FALSE" = "steelblue", "TRUE" = "red"),
                     labels = c("Real", "Junk")) +
  labs(x = "Lambda", y = "Coefficient (B)", color = "Predictor type",
       title = "Coefficient paths across lambda") +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey50")
```

### Mu recovery across lambda

```{r mu-recovery-path, fig.width=6, fig.height=4}
rmse_mu <- sapply(path_fit$fits, function(fit) {
  metric_rmse(mu_true, predict(fit, dat_test, type = "mu"))
})

plot(path_fit$summary$lambda, rmse_mu, type = "b", pch = 19,
     xlab = "Lambda", ylab = "rmse(true mu, fitted mu)",
     main = "Mean prediction recovery across lambda path")
```

## 5. Inner Cross-Validation

Note the outer cross validation (below), has this part built into
it, but it's shown here to illustrate. 
Use environmentally-blocked folds to select the best lambda via inner CV.

Note--here the 'inner' cross validation is being done on the entire dataset
so it leads to a smaller lambda then the outer cv (i.e. bigger dataset then when 
done inside of the outer CV). 

```{r inner-cv}
# create environmental clusters for fold construction
# use only real predictors (not junk) for defining env space
real_pred_vars <- c("tmean", "ppt", "vpd", "sand")
clust <- make_env_clusters(
  data = dat,
  vars = real_pred_vars,
  k = 20,
  seed = 1
)
dat$env_cluster <- clust$env_cluster

inner_folds <- make_cluster_folds(
  env_cluster = dat$env_cluster,
  n_folds = 3,
  seed = 1
)


inner_cv <- run_inner_cv(
  data = dat,
  folds = inner_folds,
  en_alpha = en_alpha,
  lambda_max_fun = cwexp_lambda_max_l1_tmb,
  lambda_max_args = list(
    formula = formula_full,
    cover_cols = cover_cols,
    dll_en = dll_en
  ),
  lambda_path_args = list(
    n_lambda = 10,
    include_zero = TRUE
  ),
  fit_path_fun = cwexp_fit_lambda_path_tmb,
  fit_path_args = list(
    formula = formula_full,
    cover_cols = cover_cols,
    dll = dll_en,
    include_report = FALSE
  ),
  keep_fold_results = TRUE
)


cat("Selected lambda (inner CV):", inner_cv$selected$lambda, "\n")
```

### Score vs lambda (averaged across inner folds)

```{r inner-cv-scores, fig.width=7, fig.height=4}
score_summary <- inner_cv$score_summary

ggplot(score_summary, aes(x = lambda, y = mae_log)) +
  geom_line() +
  geom_point() +
  geom_errorbar(aes(ymin = mae_log - mae_log_se,
                    ymax = mae_log + mae_log_se),
                width = 0) +
  geom_vline(xintercept = inner_cv$selected$lambda,
             linetype = "dashed", color = "red") +
  labs(x = "Lambda", y = "MAE (log scale)",
       title = "Inner CV: holdout MAE vs lambda",
       subtitle = paste("Selected lambda =",
                        round(inner_cv$selected$lambda, 4)))
```

## 6. Outer Cross-Validation

Nested CV to get an honest estimate of out-of-sample performance.

```{r outer-cv}

outer_folds <- make_cluster_folds(
  env_cluster = dat$env_cluster,
  n_folds = 3,
  seed = 2
)
  
outer_cv <- run_outer_cv(
  data = dat,
  outer_folds = outer_folds,
  en_alpha = en_alpha, 
  lambda_max_args = list(
    formula = formula_full,
    cover_cols = cover_cols,
    dll_en = dll_en
  ),
  lambda_path_args = list(
    n_lambda = 10,
    include_zero = TRUE
  ),
  fit_path_args = list(
    formula = formula_full,
    cover_cols = cover_cols,
    dll = dll_en
  ),
  final_fit_args = list(
    formula = formula_full,
    cover_cols = cover_cols,
    dll = dll_en,
    include_report = FALSE
  ),
  final_score_args = list(
    metrics = c("mae_log", "rmse_log"),
    predict_type = "mu"
  ),
  keep_inner_cv = TRUE
)

```

```{r outer-cv-results}
cat("Lambda selected per outer fold:\n")
print(outer_cv$outer_scores[, c("outer_fold_id", "selected_lambda",
                                 "mae_log", "rmse_log")])
cat("\nFinal lambda (median across folds):", outer_cv$lambda_final, "\n")
```

## 7. Final Model

Refit on all data at the final lambda.

```{r final-fit}

final_fit <- cwexp_fit_tmb(
  data = dat,
  formula = formula_full,
  cover_cols = cover_cols,
  dll = dll_en,
  en_alpha = en_alpha,
  lambda = outer_cv$lambda_final,
  penalty = 'elastic_net'
)

mu_hat_final <- predict(final_fit, dat, type = "mu")

cat("Final model:\n")
cat("  cor(true mu, fitted mu):", round(cor(mu_true, mu_hat_final), 3), "\n")
cat("  Estimated sigma:", round(final_fit$par$sigma, 3),
    " (true:", dummy$par$sigma, ")\n")
cat("  Lambda used:", outer_cv$lambda_final, "\n")
```

## Final model--comparisons to truth

### Compare final B to truth

```{r compare-B-final, fig.width=6, fig.height=5}
B_hat_final_vec <- as.numeric(final_fit$par$B)

lims <- range(c(B_true_vec, B_hat_final_vec))
plot(B_true_vec, B_hat_final_vec, pch = 16,
     col = ifelse(is_junk, "red", "black"),
     xlab = "True B", ylab = "Estimated B (final, regularized)",
     main = "B recovery: regularized final model",
     ylim = lims, xlim = lims)
abline(0, 1, col = "grey50", lty = 2)
abline(h = 0, v = 0, col = "grey80")
legend("topleft", legend = c("Real predictors", "Junk predictors"),
       col = c("black", "red"), pch = 16, bty = "n")
```

### Compare final B to unregularized estimate

```{r compare-B-unreg, fig.width=6, fig.height=5}
B_hat_fit0_vec <- as.numeric(fit0$par$B)

lims <- range(c(B_hat_fit0_vec, B_hat_final_vec))
plot(B_hat_fit0_vec, B_hat_final_vec, pch = 16,
     col = ifelse(is_junk, "red", "black"),
     xlab = "Estimated B (unregularized)", ylab = "Estimated B (final, regularized)",
     main = "B: regularized final model vs unregularized model",
     ylim = lims, xlim = lims)
abline(0, 1, col = "grey50", lty = 2)
abline(h = 0, v = 0, col = "grey80")
legend("topleft", legend = c("Real predictors", "Junk predictors"),
       col = c("black", "red"), pch = 16, bty = "n")
```


### Junk predictor shrinkage

Did the elastic net push junk coefficients toward zero?

```{r junk-shrinkage}
junk_rows <- grep("^junk", rownames(dummy$par$B))
B_junk_unpen <- fit0$par$B[junk_rows, ]
B_junk_final <- final_fit$par$B[junk_rows, ]

cat("Max |B_junk| unpenalized:", round(max(abs(B_junk_unpen)), 4), "\n")
cat("Max |B_junk| regularized:", round(max(abs(B_junk_final)), 4), "\n")
```

### Predicted mean (mu) biomass of each group (vs truth)


```{r}
# table
compare_group_to_truth(
  truth = dummy,
  newdata = dat_test,
  fits = list(unpenalized = fit0, regularized = final_fit)
) |>
  dplyr::mutate(dplyr::across(where(is.numeric), \(x) round(x, 3))) |>
  knitr::kable()

# plot

plot_group_recovery(truth = dummy, fit = fit0,
                    newdata = dat_test) +
  labs(subtitle = 'Unpenalized model')
plot_group_recovery(truth = dummy, fit = final_fit,
                    newdata = dat_test) +
  labs(subtitle = 'Regularized model')
```
### biomass at 100%: Predicted vs true


```{r}

plot_group_recovery(truth = dummy, fit = fit0,
                    newdata = dat_test,
                    weighted = FALSE) +
  labs(subtitle = 'Unpenalized model')
plot_group_recovery(truth = dummy, fit = final_fit,
                    newdata = dat_test,
                    weighted = FALSE) +
  labs(subtitle = 'Regularized model')
```

### General summary stats Against Truth 

Since we know the true mu for our simulated data, we can directly compare
how well the unpenalized vs regularized model recovers it.

```{r truth-comparison}
metrics <- compare_to_truth(truth = dummy,
                            newdata = dat_test,
                            fits = list(unpenalized = fit0, regularized = final_fit))
                            
knitr::kable(metrics, digits = 2)
```

## Summary

This vignette demonstrated the full cwexp pipeline on simulated data:

1. **Simulated data** with known parameters, including junk predictors.
2. **Unpenalized fit** as a baseline — recovers the general shape but
   overfits on junk predictors.
3. **Lambda path** showing coefficient shrinkage and prediction accuracy
   as a function of regularization strength.
4. **Inner CV** to select lambda using environmentally-blocked folds.
5. **Outer CV** for honest out-of-sample performance estimation.
6. **Final model** refit at the selected lambda, with improved junk
   predictor shrinkage and comparable or better recovery of the true mean.
