---
title: "Climate PCA: variable selection"
author: "Martin Holdrege"
date: "`r Sys.Date()`"
params:
  data_version: "d05"
  model_type: "herb"   # which sample to use ("herb" or "woody")
  n_top: 5             # top N variables per PC to report
  test_run: FALSE
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

## Purpose

Run PCA on the climate predictor variables to inform variable selection.
Goals:

1. Show the % variance explained per principal component (scree plot) to
   judge how many axes capture most of the climate variability.
2. Identify the variables loading most strongly on each PC, so a small
   subset (~4-5) of interpretable variables can be chosen that together
   represent most of the variation across CONUS climate.

## Setup

```{r source-and-load}
source("Functions/init.R")
source_functions()
theme_set(theme_classic())

vd <- params$data_version
model_type <- params$model_type
n_top <- params$n_top
test_run <- params$test_run

# load the sampled training data
p_dat <- file.path(
  paths$large,
  "Data_processed/BiomassQuantityData/analysis_ready",
  paste0("biomass_", model_type, "_sample_", vd, ".rds")
)
stopifnot(file.exists(p_dat))
dat_obj <- readRDS(p_dat)
dat <- dat_obj$data

if(test_run) {
  dat <- dplyr::sample_n(dat, size = 1e3)
}

cat("Loaded:", p_dat, "\n")
cat("Rows:", nrow(dat), " Cols:", ncol(dat), "\n")
```

## Select climate variables

Drop coordinates, response, cover, and region columns. Everything else is
treated as a climate variable for the PCA. `prcomp(scale. = TRUE)` re-scales
all variables, so it doesn't matter whether they were pre-z-scored.

```{r select-vars}
non_climate <- c("x", "y", "totalBio", "totalHerbaceousCov",
                 "totalTreeCov", "totalShrubCov", "region", "awc")

clim_vars <- setdiff(names(dat), non_climate)
clim_vars <- clim_vars[sapply(dat[clim_vars], is.numeric)]

cat("Climate variables (", length(clim_vars), "):\n", sep = "")
cat(paste(clim_vars, collapse = ", "), "\n")

dat_clim <- dat[, clim_vars, drop = FALSE]

# drop incomplete rows
n_before <- nrow(dat_clim)
dat_clim <- dat_clim[complete.cases(dat_clim), , drop = FALSE]
n_after <- nrow(dat_clim)
cat("Dropped", n_before - n_after, "rows with NAs (",
    n_after, "remaining)\n")
```

## Run PCA

```{r run-pca}
pca <- prcomp(dat_clim, scale. = TRUE, center = TRUE)

# variance summary
var_df <- tibble(
  PC = factor(paste0("PC", seq_along(pca$sdev)),
              levels = paste0("PC", seq_along(pca$sdev))),
  pc_num = seq_along(pca$sdev),
  variance = pca$sdev^2,
  prop_var = variance / sum(variance),
  cum_var = cumsum(prop_var)
)
```

## Scree plot

Bars show variance explained by each PC; line shows cumulative variance.
Horizontal reference at 90%.

```{r scree, fig.width = 7, fig.height = 4}
ggplot(var_df, aes(x = pc_num)) +
  geom_col(aes(y = prop_var), fill = "steelblue", alpha = 0.7) +
  geom_line(aes(y = cum_var), color = "black") +
  geom_point(aes(y = cum_var), color = "black", size = 1.5) +
  geom_hline(yintercept = 0.9, linetype = "dashed", color = "grey40") +
  scale_x_continuous(breaks = var_df$pc_num) +
  scale_y_continuous(
    labels = scales::percent,
    sec.axis = sec_axis(~ ., name = "Cumulative variance",
                        labels = scales::percent)
  ) +
  labs(x = "Principal component",
       y = "Variance explained",
       title = "Scree plot")
```

```{r var-table}
var_df |>
  mutate(across(c(prop_var, cum_var), ~ scales::percent(.x, accuracy = 0.1))) |>
  select(PC, variance, prop_var, cum_var) |>
  knitr::kable(caption = "Variance explained by each PC",
               col.names = c("PC", "Variance", "% variance", "Cumulative %"),
               digits = 3)
```

## Loadings

Loadings (rotation matrix) for all PCs. Variables are ordered by
hierarchical clustering on the loadings, so variables with similar loading
patterns sit next to each other -- these are the substitutable variables.

```{r prep-loadings}
loadings_mat <- pca$rotation

# order variables by hierarchical clustering on loadings
hc <- hclust(dist(loadings_mat))
var_order <- rownames(loadings_mat)[hc$order]

loadings_long <- as.data.frame(loadings_mat) |>
  rownames_to_column("variable") |>
  pivot_longer(-variable, names_to = "PC", values_to = "loading") |>
  mutate(
    PC = factor(PC, levels = colnames(loadings_mat)),
    variable = factor(variable, levels = var_order)
  )
```

### Loadings heatmap

```{r loadings-heatmap, fig.width = 9, fig.height = 7}
max_abs <- max(abs(loadings_long$loading))

ggplot(loadings_long, aes(x = PC, y = variable, fill = loading)) +
  geom_tile(color = "white", linewidth = 0.2) +
  scale_fill_gradient2(low = "#b2182b", mid = "white", high = "#2166ac",
                       midpoint = 0, limits = c(-max_abs, max_abs)) +
  labs(x = "Principal component", y = NULL,
       title = "PCA loadings (variables clustered by loading similarity)") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        panel.grid = element_blank())
```

### Top loading variables per PC

For each PC up to 90% cumulative variance, the top `r n_top` variables by
absolute loading. Sign is preserved so direction of association is visible.

```{r top-loadings}
n_pcs_keep <- min(which(var_df$cum_var >= 0.95))

top_tbl <- loadings_long |>
  mutate(pc_num = as.integer(str_remove(PC, "PC"))) |>
  filter(pc_num <= n_pcs_keep) |>
  group_by(PC) |>
  arrange(desc(abs(loading)), .by_group = TRUE) |>
  slice_head(n = n_top) |>
  mutate(rank = row_number(),
         loading = round(loading, 3)) |>
  ungroup()

# wide format: rows = rank, cols = PC, values = "variable (loading)"
top_wide <- top_tbl |>
  mutate(entry = paste0(variable, " (", sprintf("%+.2f", loading), ")")) |>
  select(rank, PC, entry) |>
  pivot_wider(names_from = PC, values_from = entry)

knitr::kable(
  top_wide,
  caption = paste0("Top ", n_top, " variables per PC (by |loading|), ",
                   "for PCs covering >=95% cumulative variance. ",
                   "Sign indicates direction.")
)
```

## Correlations among top-loading variables

Pearson correlations among all unique variables that appeared in the
top-`r n_top` for any PC above. Variables ordered by hierarchical
clustering on the correlation matrix, so highly correlated variables
cluster together. Use this to identify substitutable variables when
choosing a small representative set.

```{r corr-heatmap, fig.width = 9, fig.height = 8}
top_vars <- unique(as.character(top_tbl$variable))

cor_mat <- cor(dat_clim[, top_vars], method = "pearson")

# order by hclust on (1 - |cor|) so similar variables cluster
hc_cor <- hclust(as.dist(1 - abs(cor_mat)))
cor_order <- rownames(cor_mat)[hc_cor$order]

cor_long <- as.data.frame(cor_mat) |>
  rownames_to_column("var1") |>
  pivot_longer(-var1, names_to = "var2", values_to = "r") |>
  mutate(
    var1 = factor(var1, levels = cor_order),
    var2 = factor(var2, levels = cor_order)
  )

ggplot(cor_long, aes(x = var1, y = var2, fill = r)) +
  geom_tile(color = "white", linewidth = 0.2) +
  geom_text(aes(label = sprintf("%.2f", r)), size = 2.7) +
  scale_fill_gradient2(low = "#b2182b", mid = "white", high = "#2166ac",
                       midpoint = 0, limits = c(-1, 1)) +
  labs(x = NULL, y = NULL,
       title = "Pearson correlations among top-loading variables") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        panel.grid = element_blank())
```

## Climate-space coverage by selected variables

For a chosen set of variables, "coverage" = mean R² across all `r length(clim_vars)`
climate variables when each is regressed on the selected set. This is the
direct analog of "first K PCs cover Y% of variance," but using
interpretable raw variables instead of PCs. Because the raw variables are
correlated, the per-step contributions depend on the order; only the
cumulative total is order-invariant.

```{r coverage-fn}
#' Mean R^2 across all climate variables given a set of predictors
#'
#' Regresses each variable in `dat` on the columns named in `predictors`
#' and averages the R^2 values. When a variable is itself in `predictors`
#' its R^2 is set to 1.
#'
#' @param dat Data frame of climate variables (all numeric).
#' @param predictors Character vector of column names to use as predictors.
#'
#' @return Numeric scalar: mean R^2.
mean_r2_coverage <- function(dat, predictors) {
  stopifnot(all(predictors %in% names(dat)))
  X <- dat[, predictors, drop = FALSE]
  r2s <- sapply(names(dat), function(v) {
    if (v %in% predictors) return(1)
    summary(lm(dat[[v]] ~ ., data = X))$r.squared
  })
  mean(r2s)
}

#' Cumulative coverage as variables are added in a specified order
#'
#' @param dat Data frame of climate variables.
#' @param ordered_vars Character vector of variable names in the order
#'   they should be added.
#'
#' @return Tibble with columns `step`, `variable`, `cum_coverage`,
#'   `marginal`.
cumulative_coverage <- function(dat, ordered_vars) {
  out <- tibble::tibble(
    step = seq_along(ordered_vars),
    variable = ordered_vars,
    cum_coverage = NA_real_
  )
  for (i in seq_along(ordered_vars)) {
    out$cum_coverage[i] <- mean_r2_coverage(dat, ordered_vars[1:i])
  }
  out$marginal <- c(out$cum_coverage[1], diff(out$cum_coverage))
  out
}


```

### Specified selection

```{r coverage-specified}
# edit to taste
selected_vars <- c("MAT", "MAP", "PrecipTempCorr", "isothermality")

stopifnot(all(selected_vars %in% names(dat_clim)))

specified_cov <- cumulative_coverage(dat_clim, selected_vars)

specified_cov |>
  mutate(across(c(cum_coverage, marginal),
                ~ scales::percent(.x, accuracy = 0.1))) |>
  knitr::kable(caption = "mean R2 when regressed against across all other climate variables",
               col.names = c("Step", "Variable", "Cumulative", "Marginal"))
```




## Session Info

```{r session-info}
cat("Data version:", vd, "\n")
cat("Model type:", model_type, "\n")
cat("Rendered:", format(Sys.time(), "%Y-%m-%d %H:%M:%S"), "\n")
sessionInfo()
```
