---
title: "Data Exploration & Purer Sample Diagnostics"
author: "Martin Holdrege"
date: "`r Sys.Date()`"
params:
  data_version: "d02"
  purer_version: "p02"
  filter_vars: !r c("precip_CLIM", "tmean_CLIM")
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.width = 10, fig.height = 6)
```

## Purpose

Explore the input data and evaluate the effect of purer pixel selection.
Raw data sections are cached to avoid redundant computation when only the
purer selection changes.

## Setup

```{r source-and-load}
source("Functions/init.R")
source_functions()

vd <- params$data_version
vp <- params$purer_version

pfts <- opt$pfts
cover_cols <- paste0(pfts, "Cov")
pft_labels <- pfts

# --- load full data ---
dat_full <- read_analysis_ready(vd = vd)

cat("Data version:", vd, "\n")
cat("N rows:", nrow(dat_full), "\n")
cat("Cover columns:", paste(cover_cols, collapse = ", "), "\n")
```

# Part 1: Raw Data {.tabset}

## 1.1 Cover Distributions

```{r cover-histograms, fig.height=8, cache=TRUE}
cover_long <- dat_full |>
  select(all_of(cover_cols)) |>
  pivot_longer(everything(), names_to = "PFT", values_to = "cover") |>
  mutate(PFT = stringr::str_remove(PFT, "Cov$"),
         PFT = factor(PFT, levels = pft_labels))

ggplot(cover_long, aes(x = cover)) +
  geom_histogram(bins = 50, fill = "steelblue", color = "white") +
  facet_wrap(~ PFT, scales = "free_y") +
  labs(x = "Cover fraction", y = "Count",
       title = "Cover distribution per PFT")
```

```{r cover-near-zero, cache=TRUE}
# fraction of pixels with very low cover per PFT
thresholds <- c(0.01, 0.05, 0.10)
near_zero <- expand.grid(PFT = pft_labels, threshold = thresholds) |>
  rowwise() |>
  mutate(
    frac_below = mean(dat_full[[paste0(PFT, "Cov")]] < threshold)*100
  ) |>
  ungroup() |>
  pivot_wider(names_from = threshold, values_from = frac_below,
              names_prefix = "below_")

knitr::kable(near_zero, digits = 3,
             caption = "%of pixels with cover below threshold")
```

## 1.2 PFT Dominance

```{r pft-dominance, cache=TRUE}
# which PFT has highest cover in each pixel?
cover_mat <- as.matrix(dat_full[, cover_cols])
colnames(cover_mat) <- pft_labels
dominant <- pft_labels[apply(cover_mat, 1, which.max)]

dom_df <- tibble(dominant_PFT = factor(dominant, levels = pft_labels))

ggplot(dom_df, aes(x = dominant_PFT)) +
  geom_bar(fill = "steelblue") +
  labs(x = "Dominant PFT", y = "Count",
       title = "Which PFT has the highest cover in each pixel?") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```



## 1.3 Tree Cover vs Other PFTs

```{r tree-vs-others, fig.height=8, cache=TRUE}
tree_cols <- cover_cols[grepl("Tree", cover_cols)]
non_tree_cols <- setdiff(cover_cols, tree_cols)

# total tree cover
dat_full$total_tree_cover <- rowSums(dat_full[, tree_cols])

tree_vs_other <- dat_full |>
  select(total_tree_cover, all_of(non_tree_cols)) |>
  pivot_longer(-total_tree_cover, names_to = "PFT", values_to = "cover") |>
  mutate(PFT = stringr::str_remove(PFT, "Cov$"))

ggplot(tree_vs_other, aes(x = total_tree_cover, y = cover)) +
  geom_hex(bins = 50) +
  scale_fill_viridis_c(trans = "log10") +
  facet_wrap(~ PFT, scales = "free_y") +
  labs(x = "Total tree cover", y = "Cover of other PFT",
       title = "Tree cover vs other PFT cover")
```

## 1.3 Tree Cover vs Other PFTs

```{r shrub-vs-others, fig.height=8, cache=TRUE}
shrub_cols <- 'shrubCov'
non_shrub_cols <- setdiff(cover_cols, shrub_cols)

# total shrub cover
dat_full$total_shrub_cover <- rowSums(dat_full[, shrub_cols])

shrub_vs_other <- dat_full |>
  select(total_shrub_cover, all_of(non_shrub_cols)) |>
  pivot_longer(-total_shrub_cover, names_to = "PFT", values_to = "cover") |>
  mutate(PFT = stringr::str_remove(PFT, "Cov$"))

ggplot(shrub_vs_other, aes(x = total_shrub_cover, y = cover)) +
  geom_hex(bins = 50) +
  scale_fill_viridis_c(trans = "log10") +
  facet_wrap(~ PFT, scales = "free_y") +
  labs(x = "Total shrub cover", y = "Cover of other PFT",
       title = "shrub cover vs other PFT cover")
```

## 1.4 Total Biomass vs Cover

```{r bio-vs-cover, fig.height=8, cache=TRUE}
bio_cover <- dat_full |>
  select(totalBio, all_of(cover_cols)) |>
  pivot_longer(-totalBio, names_to = "PFT", values_to = "cover") |>
  mutate(PFT = stringr::str_remove(PFT, "Cov$"),
         PFT = factor(PFT, levels = pft_labels))

ggplot(bio_cover, aes(x = cover, y = totalBio)) +
  geom_hex(bins = 50) +
  scale_fill_viridis_c(trans = "log10") +
  facet_wrap(~ PFT, scales = "free") +
  labs(x = "Cover fraction", y = "Total Biomass (Mg/ha)",
       title = "Total biomass vs PFT cover")
```

## 1.5 Biomass Across Climate (Quantile Plots)

```{r bio-quantiles, fig.width=10, fig.height=6, cache=TRUE}
pred_vars_explore <- c("tmean_CLIM", "precip_CLIM", "PrecipTempCorr_CLIM", "sand")

quant_bio <- predvars2deciles(
  df = dat_full,
  response_vars = "totalBio",
  pred_vars = pred_vars_explore
)

decile_dotplot(
  yvar = "totalBio",
  df = quant_bio,
  add_predicted = FALSE,
  ylab = "Total Biomass (Mg/ha)",
  title = "Observed total biomass across climate (raw data)"
)
```

## 1.6 Biomass Across Climate — Filtered

Observed biomass vs climate, split by wet/dry (or other filter variables).

```{r bio-quantiles-filtered, fig.width=10, fig.height=8, cache=TRUE}
filter_vars <- params$filter_vars

# plot biomass vs each climate var, filtered by each filter var
for (fv in filter_vars) {
  pred_vars_filt <- setdiff(pred_vars_explore, fv)
  
  quant_filt <- predvars2deciles(
    df = dat_full,
    response_vars = "totalBio",
    pred_vars = pred_vars_filt,
    filter_var = TRUE,
    filter_vars = fv
  )
  
  # simple filtered plot
  g <- ggplot(quant_filt, aes(x = mean_value, y = totalBio,
                               color = percentile_category)) +
    geom_point(size = 0.5) +
    facet_wrap(~ name, scales = "free_x") +
    labs(x = "Mean of climate quantile",
         y = "Total Biomass (Mg/ha)",
         title = paste("Total biomass across climate, filtered by", fv),
         color = paste(fv, "percentile")) +
    theme(legend.position = "top")
  
  print(g)
}
```

# Part 2: Purer Sample Diagnostics {.tabset}

```{r load-purer}
# load the fitted model object which contains purer info
# (or re-run purer selection here if you want to test new settings)
# For now, re-run purer selection with stored config

p_spec <- purer_specs[[vp]]
purer <- select_purer_by_region(
  dat = dat_full,
  cover_vars = cover_cols,
  region_col = "region",
  q = p_spec$q,
  min_raw_cover = p_spec$min_raw_cover,
  seed = 42
)

dat_purer <- purer$data

cat("Purer version:", vp, "\n")
cat("N selected:", nrow(dat_purer), "\n")
cat("N full:", nrow(dat_full), "\n")
cat("Selection rate:", round(nrow(dat_purer) / nrow(dat_full), 3), "\n")
```

## 2.1 Selection Summary

```{r purer-summary}
cat("Thresholds per PFT per region:\n")
print(purer$thresholds)

# n selected per PFT
if ("selected_for" %in% names(dat_purer)) {
  purer_by_pft <- dat_purer |>
    count(selected_for) |>
    mutate(frac = n / sum(n))
  
  knitr::kable(purer_by_pft, caption = "Pixels selected per PFT")
}
```

## 2.2 Cover Distributions: Purer vs Full

```{r cover-purer-vs-full, fig.height=8}
cover_purer_long <- dat_purer |>
  select(all_of(cover_cols)) |>
  pivot_longer(everything(), names_to = "PFT", values_to = "cover") |>
  mutate(source = "Purer",
         PFT = stringr::str_remove(PFT, "Cov$"),
         PFT = factor(PFT, levels = pft_labels))

cover_full_long <- cover_long |>
  mutate(source = "Full")

cover_compare <- bind_rows(cover_full_long, cover_purer_long)

ggplot(cover_compare, aes(x = cover, fill = source)) +
  geom_histogram(bins = 50, position = "identity", alpha = 0.5) +
  facet_wrap(~ PFT, scales = "free_y") +
  scale_fill_manual(values = c("Full" = "grey60", "Purer" = "steelblue")) +
  labs(x = "Cover fraction", y = "Count",
       title = "Cover distributions: full data vs purer sample",
       fill = "Dataset")
```

## 2.3 Non-Focal PFT Cover in Purer Subsets

For each PFT's purer subset: how much cover of other PFTs remains?
This is the key diagnostic for identifiability.

```{r nonfocal-cover, fig.height=10}
if ("selected_for" %in% names(dat_purer)) {
  
  nonfocal_list <- lapply(cover_cols, function(focal_col) {
    focal_label <- stringr::str_remove(focal_col, "Cov$")
    subset_rows <- dat_purer$selected_for == focal_col |
      dat_purer$selected_for == focal_label
    
    if (sum(subset_rows, na.rm = TRUE) == 0) return(NULL)
    
    other_cols <- setdiff(cover_cols, focal_col)
    
    dat_purer[subset_rows, other_cols, drop = FALSE] |>
      pivot_longer(everything(), names_to = "other_PFT",
                   values_to = "cover") |>
      mutate(focal_PFT = focal_label,
             other_PFT = stringr::str_remove(other_PFT, "Cov$"))
  })
  
  nonfocal_df <- bind_rows(nonfocal_list) |>
    mutate(focal_PFT = factor(focal_PFT, levels = pft_labels),
           other_PFT = factor(other_PFT, levels = pft_labels))
  
  ggplot(nonfocal_df, aes(x = other_PFT, y = cover)) +
    geom_boxplot(fill = "steelblue", alpha = 0.5, outlier.size = 0.3) +
    facet_wrap(~ focal_PFT, labeller = label_both) +
    labs(x = "Other PFT", y = "Cover in focal PFT's purer subset",
         title = "Non-focal PFT cover within each purer subset") +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
}
```

## 2.4 Climate Coverage: Purer vs Full

```{r climate-purer-vs-full, fig.height=6}
clim_vars <- c("tmean_CLIM", "precip_CLIM", "PrecipTempCorr_CLIM", "sand")
clim_vars <- clim_vars[clim_vars %in% names(dat_full)]

clim_full <- dat_full |>
  select(all_of(clim_vars)) |>
  pivot_longer(everything(), names_to = "variable", values_to = "value") |>
  mutate(source = "Full")

clim_purer <- dat_purer |>
  select(all_of(clim_vars)) |>
  pivot_longer(everything(), names_to = "variable", values_to = "value") |>
  mutate(source = "Purer")

clim_compare <- bind_rows(clim_full, clim_purer)

ggplot(clim_compare, aes(x = value, fill = source)) +
  geom_density(alpha = 0.4) +
  facet_wrap(~ variable, scales = "free") +
  scale_fill_manual(values = c("Full" = "grey60", "Purer" = "steelblue")) +
  labs(x = "Value", y = "Density",
       title = "Climate coverage: full data vs purer sample",
       fill = "Dataset")
```

## 2.5 Biomass Distribution: Purer vs Full

```{r bio-purer-vs-full}
bio_compare <- bind_rows(
  tibble(totalBio = dat_full$totalBio, source = "Full"),
  tibble(totalBio = dat_purer$totalBio, source = "Purer")
)

ggplot(bio_compare, aes(x = log(totalBio), fill = source)) +
  geom_density(alpha = 0.4) +
  scale_fill_manual(values = c("Full" = "grey60", "Purer" = "steelblue")) +
  labs(x = "log(Total Biomass)", y = "Density",
       title = "Biomass distribution: full data vs purer sample",
       fill = "Dataset")
```

## Session Info

```{r session-info}
cat("Data version:", vd, "\n")
cat("Purer version:", vp, "\n")
cat("Rendered:", format(Sys.time(), "%Y-%m-%d %H:%M:%S"), "\n")
```
