---
title: "Data Exploration & Purer Sample Diagnostics"
author: "Martin Holdrege"
date: "`r Sys.Date()`"
params:
  model_type: "woody"
  data_version: "d06"
  purer_version: "p04.2"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.width = 10, fig.height = 6)
```

## Purpose

Explore the input data and evaluate the effect of purer pixel selection
for the `r params$model_type` model.

## Setup

```{r source-and-load}
source("Functions/init.R")
source_functions()

vd <- params$data_version
vp <- params$purer_version
model_type <- params$model_type
is_woody <- model_type == "woody"

# --- load training data ---
data_obj <-read_analysis_ready_hw(vd = vd, model_type = model_type)

dat_full <- data_obj$data
cover_cols <- data_obj$cover_cols
scale_df <- data_obj$scale
pred_vars <- data_obj$pred_vars

pft_labels <- stringr::str_remove(cover_cols, "Cov$")

# unstandardize climate for interpretable plots
dat_full_orig <- unstandardize(dat_full, scale_df, vars = pred_vars)

cat("Model type:", model_type, "\n")
cat("Data version:", vd, "\n")
cat("N rows:", nrow(dat_full), "\n")
cat("Cover columns:", paste(cover_cols, collapse = ", "), "\n")
cat("Pred vars:", paste(pred_vars, collapse = ", "), "\n")
```

# Part 1: Raw Data {.tabset}

## 1.1 Cover Distributions

```{r cover-histograms, fig.height=4*ceiling(length(cover_cols)/2), cache=TRUE}
cover_long <- dat_full |>
  dplyr::select(dplyr::all_of(cover_cols)) |>
  tidyr::pivot_longer(everything(), names_to = "PFT", values_to = "cover") |>
  dplyr::mutate(PFT = stringr::str_remove(PFT, "Cov$"),
                PFT = factor(PFT, levels = pft_labels))

ggplot(cover_long, aes(x = cover)) +
  geom_histogram(bins = 50, fill = "steelblue", color = "white") +
  facet_wrap(~ PFT, scales = "free_y") +
  labs(x = "Cover (%)", y = "Count",
       title = "Cover distribution per PFT")
```

```{r cover-near-zero, cache=TRUE}
thresholds <- c(0.01, 0.05, 0.1)
near_zero <- expand.grid(PFT = pft_labels, threshold = thresholds) |>
  dplyr::rowwise() |>
  dplyr::mutate(
    frac_below = mean(dat_full[[paste0(PFT, "Cov")]] < threshold) * 100
  ) |>
  dplyr::ungroup() |>
  tidyr::pivot_wider(names_from = threshold, values_from = frac_below,
                     names_prefix = "below_")

knitr::kable(near_zero, digits = 1,
             caption = "% of pixels with cover below threshold")
```

## 1.2 PFT Dominance

```{r pft-dominance, cache=TRUE, eval=length(cover_cols) > 1}
cover_mat <- as.matrix(dat_full[, cover_cols])
colnames(cover_mat) <- pft_labels
dominant <- pft_labels[apply(cover_mat, 1, which.max)]

dom_df <- tibble::tibble(dominant_PFT = factor(dominant, levels = pft_labels))

ggplot(dom_df, aes(x = dominant_PFT)) +
  geom_bar(fill = "steelblue") +
  labs(x = "Dominant PFT", y = "Count",
       title = "Which PFT has the highest cover in each pixel?") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## 1.3 Cover Co-occurrence

```{r cover-cooccurrence, fig.height=4*length(cover_cols), cache=TRUE, eval=length(cover_cols) > 1}
plots <- lapply(cover_cols, function(focal_col) {
  focal_label <- stringr::str_remove(focal_col, "Cov$")
  other_cols <- setdiff(cover_cols, focal_col)
  
  dat_full |>
    dplyr::select(dplyr::all_of(c(focal_col, other_cols))) |>
    tidyr::pivot_longer(-dplyr::all_of(focal_col),
                        names_to = "other_PFT", values_to = "cover") |>
    dplyr::mutate(other_PFT = stringr::str_remove(other_PFT, "Cov$")) |>
    ggplot(aes(x = .data[[focal_col]], y = cover)) +
    geom_hex(bins = 50) +
    scale_fill_viridis_c(trans = "log10") +
    facet_wrap(~ other_PFT, scales = "free_y") +
    labs(x = paste(focal_label, "cover (%)"),
         y = "Other PFT cover (%)",
         title = paste(focal_label, "cover vs other PFTs"))
})

for (g in plots) print(g)
```

## 1.4 Total Biomass vs Cover

```{r bio-vs-cover, fig.height=4*ceiling(length(cover_cols)/2), cache=TRUE}
bio_cover <- dat_full |>
  dplyr::select(totalBio, dplyr::all_of(cover_cols)) |>
  tidyr::pivot_longer(-totalBio, names_to = "PFT", values_to = "cover") |>
  dplyr::mutate(PFT = stringr::str_remove(PFT, "Cov$"),
                PFT = factor(PFT, levels = pft_labels))

ggplot(bio_cover, aes(x = cover, y = totalBio)) +
  geom_hex(bins = 50) +
  scale_fill_viridis_c(trans = "log10") +
  facet_wrap(~ PFT, scales = "free") +
  labs(x = "Cover (%)", y = "Total Biomass (Mg/ha)",
       title = "Total biomass vs PFT cover")
```


## 1.7 Biomass Distribution

```{r bio-distribution, cache=TRUE}
ggplot(dat_full, aes(x = totalBio)) +
  geom_histogram(bins = 100, fill = "steelblue", color = "white") +
  labs(x = "Total Biomass (Mg/ha)", y = "Count",
       title = paste(model_type, "biomass distribution")) +
  scale_x_continuous(trans = "log1p")
```



## 1.8 Region Distribution

```{r region-distribution, eval="region" %in% names(dat_full)}
region_counts <- dat_full |>
  dplyr::count(region, sort = TRUE)

ggplot(region_counts, aes(x = factor(region), y = n)) +
  geom_col(fill = "steelblue") +
  labs(x = "Region", y = "N pixels",
       title = "Pixels per region") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, size = 6))
```

# Part 2: Purer Sample Diagnostics {.tabset}

```{r load-purer-spec}
p_spec <- purer_specs[[vp]]

# handle herb/woody sub-specs (p04+)
if (!is.null(p_spec[[model_type]])) {
  p_spec_sub <- p_spec[[model_type]]
} else {
  p_spec_sub <- p_spec
}

do_purer <- !is.null(p_spec_sub$q) && p_spec_sub$q > 0
```

```{r purer-selection, eval=do_purer}
selection <- select_training_pixels(
  dat = dat_full,
  cover_vars = cover_cols,
  purer_spec = p_spec_sub,
  region_col = "region"
)

dat_purer <- selection$data

cat("Purer version:", vp, "\n")
cat("Method:", selection$method, "\n")
cat("N selected:", nrow(dat_purer), "\n")
cat("N full:", nrow(dat_full), "\n")
cat("Selection rate:", round(nrow(dat_purer) / nrow(dat_full), 3), "\n")
```

```{r subsample-selection, eval=!do_purer}
selection <- select_training_pixels(
  dat = dat_full,
  cover_vars = cover_cols,
  purer_spec = p_spec_sub,
  region_col = "region"
)

dat_purer <- selection$data

cat("Purer version:", vp, "\n")
cat("Method:", selection$method, "\n")
cat("N selected:", nrow(dat_purer), "\n")
cat("N full:", nrow(dat_full), "\n")
```


## 2.1 Cover Distributions: Selected vs Full

```{r cover-selected-vs-full, fig.height=4*ceiling(length(cover_cols)/2)}
cover_purer_long <- dat_purer |>
  dplyr::select(dplyr::all_of(cover_cols)) |>
  tidyr::pivot_longer(everything(), names_to = "PFT", values_to = "cover") |>
  dplyr::mutate(source = "Selected",
                PFT = stringr::str_remove(PFT, "Cov$"),
                PFT = factor(PFT, levels = pft_labels))

cover_full_long <- cover_long |>
  dplyr::mutate(source = "Full")

cover_compare <- dplyr::bind_rows(cover_full_long, cover_purer_long)

ggplot(cover_compare, aes(x = cover, fill = source)) +
  geom_histogram(aes(y = after_stat(density)),bins = 50, position = "identity", alpha = 0.5) +
  facet_wrap(~ PFT, scales = "free_y") +
  scale_fill_manual(values = c("Full" = "grey60", "Selected" = "steelblue")) +
  labs(x = "Cover", y = "Density",
       title = "Cover distributions: full data vs selected sample",
       fill = "Dataset")
```

## 2.2 Non-Focal PFT Cover in Purer Subsets

For each PFT's purer subset: how much cover of other PFTs remains?
This is the key diagnostic for identifiability.

```{r nonfocal-cover, fig.height=6, eval=do_purer && length(cover_cols) > 1 && "selected_for" %in% names(dat_purer)}
nonfocal_list <- lapply(cover_cols, function(focal_col) {
  focal_label <- stringr::str_remove(focal_col, "Cov$")
  # selected_for may contain the full col name or just the label
  subset_rows <- grepl(focal_col, dat_purer$selected_for, fixed = TRUE) |
    grepl(focal_label, dat_purer$selected_for, fixed = TRUE)
  
  if (sum(subset_rows, na.rm = TRUE) == 0) return(NULL)
  
  other_cols <- setdiff(cover_cols, focal_col)
  
  dat_purer[subset_rows, other_cols, drop = FALSE] |>
    tidyr::pivot_longer(everything(), names_to = "other_PFT",
                        values_to = "cover") |>
    dplyr::mutate(focal_PFT = focal_label,
                  other_PFT = stringr::str_remove(other_PFT, "Cov$"))
})

nonfocal_df <- dplyr::bind_rows(nonfocal_list) |>
  dplyr::mutate(focal_PFT = factor(focal_PFT, levels = pft_labels),
                other_PFT = factor(other_PFT, levels = pft_labels))

ggplot(nonfocal_df, aes(x = other_PFT, y = cover)) +
  geom_boxplot(fill = "steelblue", alpha = 0.5, outlier.size = 0.3) +
  facet_wrap(~ focal_PFT, labeller = label_both) +
  labs(x = "Other PFT", y = "Cover in focal PFT's purer subset (%)",
       title = "Non-focal PFT cover within each purer subset") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## 2.4 Climate Coverage: Selected vs Full

```{r}
quant_vars <- intersect(
  c("MAT", "MAP", "PrecipTempCorr", "isothermality", "awc"),
  pred_vars
)

dat_purer_orig <- unstandardize(dat_purer, scale_df, vars = pred_vars)

quant_full <- predvars2deciles(
  df = dat_full_orig,
  response_vars = "totalBio",
  pred_vars = quant_vars
) |> dplyr::mutate(source = "Full")

quant_purer <- predvars2deciles(
  df = dat_purer_orig,
  response_vars = "totalBio",
  pred_vars = quant_vars
) |> dplyr::mutate(source = "Selected")

quant_compare <- dplyr::bind_rows(quant_full, quant_purer)

ggplot(quant_compare, aes(x = mean_value, y = totalBio,
                           color = source, shape = source)) +
  geom_point(size = 0.75, alpha = 0.7) +
  facet_wrap(~ name, scales = "free_x") +
  scale_color_manual(values = c("Full" = "black", "Selected" = "blue")) +
  scale_shape_manual(values = c("Full" = 16, "Selected" = 17)) +
  labs(x = "Mean of predictor quantile",
       y = "Total Biomass (Mg/ha)",
       title = paste("Biomass across climate: full vs selected (", model_type, ")"),
       color = NULL, shape = NULL) +
  theme(legend.position = "top")
```


```{r climate-selected-vs-full, fig.height=6}
dat_purer_orig <- unstandardize(dat_purer, scale_df, vars = pred_vars)

clim_vars_plot <- intersect(quant_vars, pred_vars)

clim_full <- dat_full_orig |>
  dplyr::select(dplyr::all_of(clim_vars_plot)) |>
  tidyr::pivot_longer(everything(), names_to = "variable", values_to = "value") |>
  dplyr::mutate(source = "Full")

clim_purer <- dat_purer_orig |>
  dplyr::select(dplyr::all_of(clim_vars_plot)) |>
  tidyr::pivot_longer(everything(), names_to = "variable", values_to = "value") |>
  dplyr::mutate(source = "Selected")

clim_compare <- dplyr::bind_rows(clim_full, clim_purer)

ggplot(clim_compare, aes(x = value, fill = source)) +
  geom_density(alpha = 0.4) +
  facet_wrap(~ variable, scales = "free") +
  scale_fill_manual(values = c("Full" = "black", "Selected" = "blue")) +
  labs(x = "Value", y = "Density",
       title = "Climate coverage: full data vs selected sample",
       fill = "Dataset")
```



## 2.5 Biomass Distribution: Selected vs Full

```{r bio-selected-vs-full}
bio_compare <- dplyr::bind_rows(
  tibble::tibble(totalBio = dat_full$totalBio, source = "Full"),
  tibble::tibble(totalBio = dat_purer$totalBio, source = "Selected")
)

ggplot(bio_compare, aes(x = log1p(totalBio), fill = source)) +
  geom_density(alpha = 0.4) +
  scale_fill_manual(values = c("Full" = "black", "Selected" = "blue")) +
  labs(x = "log1p(Total Biomass)", y = "Density",
       title = "Biomass distribution: full data vs selected sample",
       fill = "Dataset")

ggplot(bio_compare, aes(x = totalBio, fill = source)) +
  geom_density(alpha = 0.4) +
  scale_fill_manual(values = c("Full" = "black", "Selected" = "blue")) +
  labs(x = "Total Biomass", y = "Density",
       title = "Biomass distribution: full data vs selected sample",
       fill = "Dataset")
```

## Session Info

```{r session-info}
cat("Model type:", model_type, "\n")
cat("Data version:", vd, "\n")
cat("Purer version:", vp, "\n")
cat("Rendered:", format(Sys.time(), "%Y-%m-%d %H:%M:%S"), "\n")
```
