---
title: "Classification model diagnostics"
author: "Martin Holdrege"
date: "`r Sys.Date()`"
params:
  cover_type: "classification"
  cover_model: null   # null uses opt$cover_model
  vc: null            # null uses opt$vc
  vmc: null           # null uses opt$vmc
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# Setup

```{r setup-data}
source("Functions/init.R")
source_functions()
library(glmnet)
theme_set(theme_custom1())

cover_type  <- params$cover_type
cover_model <- if (is.null(params$cover_model)) opt$cover_model else params$cover_model
vc          <- if (is.null(params$vc)) opt$vc else params$vc
vmc         <- if (is.null(params$vmc)) opt$vmc else params$vmc

fit  <- read_cover_model(cover_type, cover_model, vc, vmc)
if(is.null(fit$config$spec$engine)) fit$config$spec$engine <- 'glmnet' # for legacy reasons
spec <- fit$config$spec

# "glmnet" or "ranger"; the fit summary and a couple of labels differ

engine <- spec$engine
# for ranger, pred_in holds out-of-bag rather than in-sample predictions
in_label <- if (engine == "ranger") "out_of_bag" else "in_sample"
in_title <- if (engine == "ranger") {
  "Out-of-bag"
} else {
  "In-sample (model fit to all data)"
}

# binomial models predict a probability of forest; continuous ones (family
# gaussian, poisson or quasibinomial) predict tree cover (%), cut into classes
# the same way
score_layer <- score_name(fit)  # "prob" or "cover"
continuous <- score_layer == "cover"
score_label <- if (continuous) {
  "Predicted tree cover (%)"
} else {
  "Predicted probability of forest"
}
score_limits <- if (continuous) c(0, 100) else c(0, 1)
class_label <- "forest"  # name of the positive class, for outcome labels

# predictors in original units (log1p_MAP is plotted as MAP)
source_vars <- unique(str_remove(spec$pred_vars, "^log1p_"))

# training data joined to the saved predictions; observed class rebuilt from
# the spec
dat <- read_cover_training(vc) |>
  inner_join(fit$predictions, by = c("cell", "year")) |>
  mutate(cover_obs = .data[[spec$response]],
         observed = as.integer(cover_obs > spec$cover_threshold))
stopifnot(nrow(dat) == nrow(fit$predictions))

scenarios <- c("current", "BNU-ESM", "IPSL-CM5A-MR")
pred_dir <- file.path(paths$large, "Data_processed", "CoverData", "Predictions")
```

# Model info 

Model `r cover_type`/`r cover_model`, data `r vc`, version `r vmc`, engine
`r engine`, family `r if (continuous) spec$family else "binomial"`:
`r nrow(dat)` pixel-years; cutoff `r round(fit$threshold, 3)`
`r if (continuous) "% predicted cover (observed classes split at 10%)" else ""`

```{r info}
print(spec)
```

# Fit

## Fitted model

For glmnet, the cross-validation curve (binomial deviance vs log(lambda);
dashed lines mark min and 1se) and the non-zero coefficients at the selected
lambda. For ranger, the forest settings and the out-of-bag error. For grpreg
(hierarchical group lasso), the cross-validation curve (mean out-of-fold
deviance +/- 1 SE; dashed line at the selected lambda) and the non-zero
coefficients, each summed over the groups it was fit
in. Squares only appear with their main effect, and interactions with
both of their terms.
for these binomial models (where it's succeses out of 1 try) binomial deviance is just -2*log likelihood.
For quasibinomial glmnet models the CV curve is the binomial deviance of
cover as a proportion (cover / 100), not of the class. For gaussian glmnet
models it is the mean squared error of cover on the scale it was fit on
(e.g. log1p cover).

```{r fit-summary}
if (engine == "glmnet") {
  
  plot(fit$fit)
  
coefs <- coef(fit$fit, s = fit$lambda)
tibble(term = rownames(coefs), coefficient = as.numeric(coefs)) |>
  filter(coefficient != 0) |>
  arrange(desc(abs(coefficient))) |>
    knitr::kable(digits = 3) |>
    print()
  
} else if (engine == "ranger") {
  
  # selected fields only: print(fit$fit) deparses fit$call, which can hold
  # the whole training matrix. prediction.error is the out-of-bag Brier score.
  # Add importance = "impurity" to spec$ranger for variable importance.
  fit$fit[c("treetype", "num.trees", "mtry", "min.node.size", "splitrule",
            "num.samples", "num.independent.variables",
            "prediction.error")] |>
    map_chr(\(v) format(v, digits = 3)) |>
    enframe(name = "setting", value = "value") |>
    knitr::kable() |>
    print()
  
} else if (engine == "grpreg") {
  
  cv_plot <- ggplot(fit$fit$cv, aes(x = log(lambda), y = cvm)) +
    geom_pointrange(aes(ymin = cvm - cvsd, ymax = cvm + cvsd), size = 0.2) +
    geom_vline(xintercept = log(fit$lambda), linetype = 2) +
    labs(y = paste0("Out-of-fold ", spec$family, " deviance"))
  print(cv_plot)
  
  # coefficients on the standardized scale of the design matrix
  tibble(term = names(fit$fit$coef), coefficient = fit$fit$coef) |>
    filter(coefficient != 0) |>
    arrange(desc(abs(coefficient))) |>
    knitr::kable(digits = 3) |>
    print()
  
} else if (engine == "gam") {
  
  # CV curve over the multiplier on REML's smoothing parameters (x1 is
  # REML's own choice; larger is smoother)
  cv_plot <- ggplot(fit$fit$cv, aes(x = log10(lambda), y = cvm)) +
    geom_pointrange(aes(ymin = cvm - cvsd, ymax = cvm + cvsd), size = 0.2) +
    geom_vline(xintercept = log10(fit$lambda), linetype = 2) +
    labs(x = "log10(multiplier on REML smoothing parameters)",
         y = paste0("Out-of-fold ", spec$gam$metric))
  print(cv_plot)
  
  # effective degrees of freedom per smooth: ~1 near linear, ~0 dropped out
  fit$fit$edf |>
    arrange(desc(edf)) |>
    knitr::kable(digits = 2) |>
    print()
  
}
```

## Classification performance

AUC and the confusion matrix at the cutoff. Out-of-fold predictions come from
the environmental folds; for glmnet and grpreg those are the same folds that
chose lambda, so they are mildly optimistic. The other set is in-sample for
glmnet and grpreg, and out-of-bag for ranger.

```{r performance}

perf <- function(pred) {
  # PresenceAbsence wants the cutoff in 0-1, so predicted cover (%) goes on
  # the 0-1 scale
  cutoff <- fit$threshold
  if (continuous) {
    pred <- pred / 100
    cutoff <- cutoff / 100
  }
  pa <- data.frame(id = seq_len(nrow(dat)), obs = dat$observed, pred = pred)
  cm <- PresenceAbsence::cmx(pa, threshold = cutoff) # confusion matrix
  tibble(
    #the probability that a randomly chosen forest pixel gets a higher p̂ than a
    #randomly chosen non-forest pixel. 0.5 is random and 1 is perfect ranking.
          AUC = PresenceAbsence::auc(pa, st.dev = FALSE),
         PCC = PresenceAbsence::pcc(cm, st.dev = FALSE), # percent correctly classified
         # Of the pixels that are actually forest, it's the fraction called forest.
         sensitivity = PresenceAbsence::sensitivity(cm, st.dev = FALSE),
         # Of the pixels that are actually non-forest, it's the fraction called non-forest.
         specificity = PresenceAbsence::specificity(cm, st.dev = FALSE),
         observed_prevalence = mean(dat$observed),
         predicted_prevalence = mean(pred > cutoff))
  
}

bind_rows(perf(dat$pred_in), perf(dat$pred_oof)) |>
  mutate(predictions = c(in_label, "out_of_fold"), .before = 1) |>
  knitr::kable(digits = 3)

# confusion matrix (counts of pixel-years), out-of-fold
table(observed = dat$observed,
      predicted = as.integer(dat$pred_oof > fit$threshold)) |>
  knitr::kable(caption = "Out-of-fold confusion matrix (0 = not forest, 1 = forest)")
```

## Predicted vs observed cover

Continuous models only. Out-of-fold predicted against observed tree cover
(log1p axes; predictions outside 0-100% are drawn at the limits). Dashed
lines: the 10% threshold (vertical) and the cutoff on predicted cover
(horizontal), so the top-left corner is false forest and the bottom-right
missed forest.

```{r cover-fit, eval = continuous, fig.width = 7, fig.height = 5}
ggplot(dat, aes(x = cover_obs, y = pmin(pmax(pred_oof, 0), 100))) +
  geom_bin2d(bins = 60) +
  geom_abline(colour = "grey40") +
  geom_vline(xintercept = spec$cover_threshold, linetype = 2) +
  geom_hline(yintercept = fit$threshold, linetype = 2) +
  scale_x_continuous(trans = "log1p", breaks = c(0, 1, 5, 10, 25, 50, 100)) +
  scale_y_continuous(trans = "log1p", breaks = c(0, 1, 5, 10, 25, 50, 100)) +
  scale_fill_viridis_c(trans = "log10", name = "Pixel-years") +
  labs(x = "Observed tree cover (%)",
       y = "Out-of-fold predicted tree cover (%)")
```

# Quantile plots

Points: proportion of pixel-years that are forest, observed and predicted
(predicted class = prediction above the cutoff), within quantiles of each
predictor. Both are classes, so they're directly comparable.

```{r quantile-data}
# decile_dotplot() expects the predicted column to be <yvar>_pred
quant_df <- function(pred_col) {
  dat |>
    mutate(forest = observed,
           forest_pred = as.integer(.data[[pred_col]] > fit$threshold)) |>
    predvars2deciles(response_vars = c("forest", "forest_pred"),
                     pred_vars = source_vars)
}
quant_in  <- quant_df("pred_in")
quant_oof <- quant_df("pred_oof")
```

```{r quantile-plots, fig.width = 10, fig.height = 7}
decile_dotplot("forest", quant_in, ylab = "Proportion forest",
               add_predicted = TRUE, title = in_title,
               add_ribbon = FALSE)
decile_dotplot("forest", quant_oof, ylab = "Proportion forest",
               add_predicted = TRUE, title = "Out-of-fold",
               add_ribbon = FALSE)
```

## Filtered quantile plots

Only pixel-years in the lower and upper 20% of each filter variable, to see
whether the fit holds at the climate extremes.

```{r quantile-filtered, fig.width = 10, fig.height = 9}
filtered <- function(pred_col, title) {
  dat |>
  mutate(forest = observed,
         forest_pred = as.integer(.data[[pred_col]] > fit$threshold)) |>
  predvars2deciles(response_vars = c("forest", "forest_pred"),
                   pred_vars = source_vars, filter_var = TRUE,
                   filter_vars = source_vars) |>
  decile_dotplot_filtered("forest", df = _, ylab = "Proportion forest",
                          title = title)
}

filtered('pred_in', title = paste0('filtered (', in_title, ')'))
filtered('pred_oof', title = 'filtered (out of fold)')

```

# Partial dependence

Mean prediction (probability, or % cover for continuous models) as each
predictor varies, others at their observed values.

```{r pdp, fig.width = 10, fig.height = 6}
pdp_classification(fit, dat) |>
  plot_pdp(ylab = score_label)
```

# Accumulated local effects

Change in predicted probability as each predictor increases, relative to
the average prediction (Apley & Zhu 2020). Built from small steps within
quantile bins, using only pixels near each value, so unlike the partial
dependence plots it does not evaluate the model at combinations of
predictors that don't occur. Where the two disagree, the PDP is being
driven by those combinations. Middle 98% of each predictor. Points are
bin edges; each segment is the mean change in probability across a bin.

```{r ale, fig.width = 10, fig.height = 6}
ale_classification(fit, dat) |>
  plot_ale(ylab = if (continuous) {
    "ALE (change in predicted cover, %)"
  } else {
    "ALE (change in predicted probability)"
  })
```

## Filtered accumulated local effects

ALE computed separately within the lowest and highest 20% of each filter
variable, as in the filtered quantile plots. Each line is the subset's mean
predicted probability plus its ALE, so the lines differ in level where the
subsets differ, and in shape where the predictor interacts with the filter
variable in the fitted model.

```{r ale-filtered, fig.width = 10, fig.height = 9}
ale_classification_filtered(fit, dat) |>
  plot_ale_filtered(ylab = if (continuous) {
    "Cover scale, % (subset mean + ALE)"
  } else {
    "Probability scale (subset mean + ALE)"
  })
```

# Maps

## Misclassification at training pixels

Observed class vs predicted class, for both sets of predictions. One point
per pixel, using its most recent year.

```{r map-misclass, fig.width = 9, fig.height = 6}
plot_misclass <- function(pred_col, title) {
  misclass_sf <- dat |>
    slice_max(year, n = 1, by = cell) |>
    classify_outcome(pred_col, threshold = fit$threshold,
                     class_label = class_label) |>
    sf::st_as_sf(coords = c("x", "y"), crs = terra::crs(read_mask()))
  
  plot_points_conus(misclass_sf, color_var = "outcome", point_size = 0.3,
                    title = title,
                    colorscale = scale_colour_manual(
                      name = NULL,
                      values = outcome_colours(class_label))) +
    # larger points in the legend only
    guides(colour = guide_legend(override.aes = list(size = 3)))
}

plot_misclass("pred_in", in_title)
plot_misclass("pred_oof", "Out-of-fold")
```

## Predictor distributions by classification outcome

Density of each predictor (original units, e.g. MAP rather than log1p_MAP)
for the four outcomes in the map above, over all pixel-years. Each line
integrates to 1, so rare outcomes (e.g. missed forest) are as visible as
common ones; the lines show where along a predictor each outcome occurs, not
how many pixel-years it has.

```{r density-misclass, fig.width = 10, fig.height = 7}
dat |>
  classify_outcome("pred_in", threshold = fit$threshold,
                   class_label = class_label) |>
  plot_outcome_density(source_vars, title = in_title)
dat |>
  classify_outcome("pred_oof", threshold = fit$threshold,
                   class_label = class_label) |>
  plot_outcome_density(source_vars, title = "Out-of-fold")
```

## Out-of-fold residuals at training pixels

Mean over years of (prediction - observation) per pixel: predicted
probability - observed class, or for continuous models predicted -
observed % cover (colours capped at +/- 30). Positive: predicted more forest
than observed. For models fit to log1p(cover), predicted cover is the
back-transformed mean of log1p(cover), which runs below mean cover, so these
residuals lean negative by construction; the classification is unaffected,
because the cutoff is set on the same predictions. Quasibinomial models
predict mean cover directly, so this bias doesn't apply to them.

```{r map-residuals, fig.width = 9, fig.height = 6}
resid_sf <- dat |>
  mutate(resid = if (continuous) pred_oof - cover_obs else pred_oof - observed) |>
  summarise(resid = mean(resid), .by = c(cell, x, y)) |>
  sf::st_as_sf(coords = c("x", "y"), crs = terra::crs(read_mask()))

plot_points_conus(resid_sf, color_var = "resid", point_size = 0.3,
                  colorscale = scale_colour_gradient2(
                    name = if (continuous) {
                      "Predicted -\nobserved\n(% cover)"
                    } else {
                      "Predicted -\nobserved"
                    },
                    low = "#b2182b", high = "#2166ac",
                    limits = if (continuous) c(-30, 30) else c(-1, 1),
                    oob = scales::squish))
```

## Current climate

All of CONUS (snap mask), not only the available area: predictions stand for
potential natural vegetation, and SOILWAT2 will be run CONUS-wide.

```{r map-read}
read_pred <- function(scenario) {
  terra::rast(file.path(pred_dir, paste0(cover_type, "_", cover_model, "_",
                                         vc, "-", vmc, "_", scenario,
                                         ".tif")))
}
pred <- map(scenarios, read_pred) |> set_names(scenarios)

# binary and change rasters as categories, so legends show classes rather
# than a continuous color bar
as_class_raster <- function(r) {
  r <- terra::as.int(r)
  levels(r) <- data.frame(value = c(0, 1), class = c("not forest", "forest"))
  r
}
class_scale <- scale_fill_manual(name = NULL,
                                 values = c("not forest" = "grey85",
                                            "forest" = "darkgreen"),
                                 na.translate = FALSE)

as_change_raster <- function(r) {
  r <- terra::as.int(r)
  levels(r) <- data.frame(value = c(-1, 0, 1),
                          change = c("forest lost", "no change",
                                     "forest gained"))
  r
}
change_scale <- scale_fill_manual(name = NULL,
                                  values = c("forest lost" = "orange",
                                             "no change" = "grey90",
                                             "forest gained" = "#1b7837"),
                                  na.translate = FALSE)
```

```{r raster-summary}
terra::global(pred$current[[score_layer]], c("min", "mean", "max"),
              na.rm = TRUE)
```



```{r map-current, fig.width = 9, fig.height = 6}
# continuous predictions can fall outside 0-100%; the colours are capped
plot_map_conus(pred$current[[score_layer]],
               colorscale = scale_fill_gradient2(
                    name = if (continuous) "Cover (%)" else "Probability",
                    low = "#b2182b",
                    high = "#2166ac",
                    mid = 'grey85',
                    midpoint = fit$threshold,
                    na.value = 'transparent',
                    limits = score_limits,
                    oob = scales::squish),
               title = score_label)

plot_map_conus(as_class_raster(pred$current$class),
               colorscale = class_scale,
               title = "Predicted class")
```

## Future climate

Class under each future scenario, and the change from current (+1 becomes
forest, -1 stops being forest).

```{r map-future, fig.width = 9, fig.height = 6}
walk(setdiff(scenarios, "current"), \(scenario) {
  change <- pred[[scenario]]$class - pred$current$class
  
  print(plot_map_conus(as_class_raster(pred[[scenario]]$class),
                       colorscale = class_scale,
                       title = paste("Predicted class,", scenario)))
  
  print(plot_map_conus(as_change_raster(change),
                       colorscale = change_scale,
                       title = paste("Change in class,", scenario,
                                     "vs current")))
})
```

# Session info

```{r session-info}
sessionInfo()
```