---
title: "Cover model diagnostics"
author: "Martin Holdrege"
date: "`r Sys.Date()`"
params:
  cover_type: "cover"
  cover_model: null   # null uses opt$cover_model
  vc: null            # null uses opt$vc
  vmc: null           # null uses opt$vmc
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# Setup

```{r setup-data}
source("Functions/init.R")
source_functions()
library(glmnet)
library(patchwork)
theme_set(theme_custom1())

cover_type  <- params$cover_type
cover_model <- if (is.null(params$cover_model)) opt$cover_model else params$cover_model
vc          <- if (is.null(params$vc)) opt$vc else params$vc
vmc         <- if (is.null(params$vmc)) opt$vmc else params$vmc

fit  <- read_cover_model(cover_type, cover_model, vc, vmc)
spec <- fit$config$spec
stopifnot(inherits(fit, "cover_continuous"))

engine <- spec$engine
# for ranger, pred_in holds out-of-bag rather than in-sample predictions
in_label <- if (engine == "ranger") "out_of_bag" else "in_sample"
in_title <- if (engine == "ranger") {
  "Out-of-bag"
} else {
  "In-sample"
}

# e.g. "Tree cover (%)" for cov_tree
cover_label <- paste0(str_to_sentence(str_remove(spec$response, "^cov_")),
                      " cover (%)")

# predictors in original units (log1p_MAP is plotted as MAP)
source_vars <- unique(str_remove(spec$pred_vars, "^log1p_"))

# training data joined to the saved predictions (both % cover)
dat <- read_cover_training(vc) |>
  inner_join(fit$predictions, by = c("cell", "year")) |>
  mutate(cover_obs = .data[[spec$response]])
stopifnot(nrow(dat) == nrow(fit$predictions), !anyNA(dat$cover_obs))

scenarios <- c("current", "BNU-ESM", "IPSL-CM5A-MR")
pred_dir <- file.path(paths$large, "Data_processed", "CoverData", "Predictions")

# wall-to-wall comparison with the RAP raster: only for tree cover fit to c02,
# which is a sample of that same raster (05_rap_training_sample.R)
wall <- spec$response == "cov_tree" && vc == "c02"

```

# Model info

Model `r cover_type`/`r cover_model`, data `r vc`, version `r vmc`, engine
`r engine`, family `r spec$family`: `r nrow(dat)` pixel-years. Rows fit to:
`r if (is.language(spec$rows)) deparse1(spec$rows) else spec$rows`.

```{r info}
print(spec)
```

# Fit

## Fitted model

For glmnet, the cross-validation curve and the non-zero coefficients at the
selected lambda. The CV error is the binomial deviance of cover as a
proportion (cover / 100) for quasibinomial models, or the mean squared error
of cover on the scale it was fit on for gaussian models. For ranger, the
forest settings and out-of-bag error (mean squared error on the fit scale),
and variable importance if the spec asked for it. For gam, the
cross-validation curve over the smoothing multiplier and the effective
degrees of freedom of each smooth.

```{r fit-summary}
if (engine == "glmnet") {

  plot(fit$fit)

  coefs <- coef(fit$fit, s = fit$lambda)
  tibble(term = rownames(coefs), coefficient = as.numeric(coefs)) |>
    filter(coefficient != 0) |>
    arrange(desc(abs(coefficient))) |>
    knitr::kable(digits = 3) |>
    print()

} else if (engine == "ranger") {

  # selected fields only: print(fit$fit) deparses fit$call, which can hold
  # the whole training matrix. r.squared is the out-of-bag R^2
  fit$fit[c("treetype", "num.trees", "mtry", "min.node.size", "splitrule",
            "num.samples", "num.independent.variables",
            "prediction.error", "r.squared")] |>
    map_chr(\(v) format(v, digits = 3)) |>
    enframe(name = "setting", value = "value") |>
    knitr::kable() |>
    print()

  # variable importance, if the forest was fit with it
  if (!is.null(fit$fit$variable.importance)) {
    vip_plot <- ranger::importance(fit$fit) |>
      enframe(name = "variable", value = "importance") |>
      ggplot(aes(x = importance, y = reorder(variable, importance))) +
      geom_col() +
      labs(x = paste0("Importance (", fit$fit$importance.mode, ")"), y = NULL)
    print(vip_plot)
  }

} else if (engine == "grpreg") {

  cv_plot <- ggplot(fit$fit$cv, aes(x = log(lambda), y = cvm)) +
    geom_pointrange(aes(ymin = cvm - cvsd, ymax = cvm + cvsd), size = 0.2) +
    geom_vline(xintercept = log(fit$lambda), linetype = 2) +
    labs(y = paste0("Out-of-fold ", spec$family, " deviance"))
  print(cv_plot)

  tibble(term = names(fit$fit$coef), coefficient = fit$fit$coef) |>
    filter(coefficient != 0) |>
    arrange(desc(abs(coefficient))) |>
    knitr::kable(digits = 3) |>
    print()

} else if (engine == "gam") {

  cv_plot <- ggplot(fit$fit$cv, aes(x = log10(lambda), y = cvm)) +
    geom_pointrange(aes(ymin = cvm - cvsd, ymax = cvm + cvsd), size = 0.2) +
    geom_vline(xintercept = log10(fit$lambda), linetype = 2) +
    labs(x = "log10(multiplier on REML smoothing parameters)",
         y = paste0("Out-of-fold ", spec$gam$metric))
  print(cv_plot)

  fit$fit$edf |>
    arrange(desc(edf)) |>
    knitr::kable(digits = 2) |>
    print()

}
```

## Performance

Predicted vs observed cover (%) at the training pixel-years. Out-of-fold
predictions come from the environmental folds; for glmnet, grpreg and gam
those are the same folds that chose the penalty, so they are mildly
optimistic. The other set is in-sample for glmnet, grpreg and gam, and
out-of-bag for ranger. R^2 is 1 - SSE/SST (negative when the model does worse
than the mean); bias is mean(predicted - observed).

```{r performance}
bind_rows(cover_metrics(dat$cover_obs, dat$pred_in),
          cover_metrics(dat$cover_obs, dat$pred_oof)) |>
  mutate(predictions = c(in_label, "out_of_fold"), .before = 1) |>
  knitr::kable(digits = 3)
```

## Predicted vs observed

Out-of-fold predictions. Grey line: 1:1.

```{r obs-pred, fig.width = 7, fig.height = 5}
ggplot(dat, aes(x = cover_obs, y = pred_oof)) +
  geom_bin2d(bins = 60) +
  geom_abline(colour = "grey40") +
  scale_fill_viridis_c(trans = "log10", name = "Pixel-years") +
  labs(x = paste("Observed", str_to_lower(cover_label)),
       y = paste("Out-of-fold predicted", str_to_lower(cover_label)))
```

# Quantile plots

Points: mean cover, observed and predicted, within quantiles of each
predictor.

```{r quantile-data}
# decile_dotplot() expects the predicted column to be <yvar>_pred
quant_df <- function(pred_col) {
  dat |>
    mutate(cover = cover_obs,
           cover_pred = .data[[pred_col]]) |>
    predvars2deciles(response_vars = c("cover", "cover_pred"),
                     pred_vars = source_vars)
}
quant_in  <- quant_df("pred_in")
quant_oof <- quant_df("pred_oof")
```

```{r quantile-plots, fig.width = 10, fig.height = 7}
decile_dotplot("cover", quant_in, ylab = cover_label,
               add_predicted = TRUE, title = in_title,
               add_ribbon = TRUE)
decile_dotplot("cover", quant_oof, ylab = cover_label,
               add_predicted = TRUE, title = "Out-of-fold",
               add_ribbon = TRUE)
```

## Filtered quantile plots

Only pixel-years in the lower and upper 20% of each filter variable, to see
whether the fit holds at the climate extremes.

```{r quantile-filtered, fig.width = 10, fig.height = 9}
filtered <- function(pred_col, title) {
  dat |>
    mutate(cover = cover_obs,
           cover_pred = .data[[pred_col]]) |>
    predvars2deciles(response_vars = c("cover", "cover_pred"),
                     pred_vars = source_vars, filter_var = TRUE,
                     filter_vars = source_vars) |>
    decile_dotplot_filtered("cover", df = _, ylab = cover_label,
                            title = title)
}

filtered("pred_in", title = paste0("filtered (", in_title, ")"))
filtered("pred_oof", title = "filtered (out of fold)")
```

# Partial dependence

Mean predicted cover as each predictor varies, others at their observed
values. (`pdp_classification()` and the ALE functions work for any cover
model: they predict with `predict_score()`.)

```{r pdp, fig.width = 10, fig.height = 6}
pdp_classification(fit, dat) |>
  plot_pdp(ylab = paste("Predicted", str_to_lower(cover_label)))
```

# Accumulated local effects

Change in predicted cover as each predictor increases, relative to the
average prediction (Apley & Zhu 2020), using only pixels near each value, so
unlike the partial dependence plots it does not evaluate the model at
combinations of predictors that don't occur. Middle 98% of each predictor.

```{r ale, fig.width = 10, fig.height = 6}
ale_classification(fit, dat) |>
  plot_ale(ylab = "ALE (change in predicted cover, %)")
```

## Filtered accumulated local effects

ALE within the lowest and highest 20% of each filter variable. Each line is
the subset's mean predicted cover plus its ALE.

```{r ale-filtered, fig.width = 10, fig.height = 9}
ale_classification_filtered(fit, dat) |>
  plot_ale_filtered(ylab = "Cover, % (subset mean + ALE)")
```

# Maps

## Training pixels

Observed and predicted, and their difference, at the
training pixels (mean over years per pixel). Shared colour scale for the
first two, capped at the 99th percentile of observed cover.

```{r map-training, fig.width = 12, fig.height = 8}
px <- dat |>
  summarise(observed = mean(cover_obs), predicted = mean(pred_oof),
            predicted_in = mean(pred_in),
            .by = c(cell, x, y)) |>
  mutate(resid = predicted - observed,
         resid_in = predicted_in - observed) |>
  sf::st_as_sf(coords = c("x", "y"), crs = terra::crs(read_mask()))

upper <- quantile(px$observed, 0.99, names = FALSE)
cover_scale_pts <- colorscale_cover(name = "Cover (%)", limits = c(0, upper),
                                    aesthetics = "colour")

l <- quantile(c(px$resid, px$resid_in), c(0.05, 0.95), names = FALSE) |> 
  abs() |> 
  max()
limits_resid = c(-l, l)

cover_scale_resid = scale_colour_gradient2(
                    name = "Predicted -\nobserved\n(% cover,\n5th-9th percentiles)",
                    low = "#b2182b", high = "#2166ac",
                    limits = limits_resid, oob = scales::squish)


g1 <- plot_points_conus(px, color_var = "observed", point_size = 0.3,
                  colorscale = cover_scale_pts,
                  title = paste("Observed", str_to_lower(cover_label)))
g2 <- plot_points_conus(px, color_var = "predicted_in", point_size = 0.3,
                  colorscale = cover_scale_pts,
                  title = paste(in_title, "predicted",
                                str_to_lower(cover_label)))

g3 <- plot_points_conus(px, color_var = "resid_in", point_size = 0.3,
                  colorscale = cover_scale_resid,
                  title = paste(in_title, "residuals"))

g4 <- plot_points_conus(px, color_var = "resid", point_size = 0.3,
                  colorscale = cover_scale_resid,
                  title = "Out-of-fold residuals")


(g1 + g2)/(g3 + g4) + patchwork::plot_layout(guides = 'collect')

```

## Current climate

Predictions on all of CONUS (snap mask), not only the model's domain
(`spec$rows`) or the available area.

```{r map-read}
read_pred <- function(scenario) {
  terra::rast(file.path(pred_dir, paste0(cover_type, "_", cover_model, "_",
                                         vc, "-", vmc, "_", scenario,
                                         ".tif")))[["cover"]]
}
pred <- map(scenarios, read_pred) |> set_names(scenarios)

terra::global(pred$current, c("min", "mean", "max"), na.rm = TRUE)
```

```{r map-current, fig.width = 9, fig.height = 6}
upper_r <- raster_quantile(pred$current, 0.99)
cover_scale <- colorscale_cover(name = "Cover (%)", limits = c(0, upper_r))

plot_map_conus(pred$current, colorscale = cover_scale,
               title = paste("Predicted", str_to_lower(cover_label),
                             "(current climate)"))
```

## Wall-to-wall comparison with RAP

Tree cover models fit to c02 only. The observed layer is the 2021 RAP tree
cover raster that c02 was sampled from, in the cells passing the LCMAP and
fire masks, and restricted to the model's domain (`spec$rows`, e.g. tree
cover > 10%) as in fitting. The predictions are those above (current climate,
model fit to all training data), so the training pixels are included, and
neighbouring cells are not independent of them: this shows the spatial
pattern of the errors rather than an out-of-sample test. Note--can'
show areas not in the model's domain (e.g. if model fit to tree cover < 10%)

```{r wall-data, eval = wall}
rap_file <- file.path(paths$large, "Data_processed/CoverData/rap",
                      "RAP_v3_cover_2021_1000m.tif")
obs_r <- terra::rast(rap_file)[["tree"]] |>
  align_raster(pred$current)  # GEE exports are one cell larger
names(obs_r) <- "cov_tree"
stopifnot(terra::compareGeom(obs_r, pred$current))

pred_r <- terra::mask(pred$current, obs_r)
diff_r <- pred_r - obs_r

```


```{r wall-maps, eval = wall, fig.width = 9, fig.height = 6}
upper_w <- raster_quantile(obs_r, 0.99)
wall_scale <- colorscale_cover(name = "Cover (%)", limits = c(0, upper_w))

plot_map_conus(obs_r, colorscale = wall_scale,
               title = "RAP tree cover, 2021")
plot_map_conus(pred_r, colorscale = wall_scale,
               title = "Predicted tree cover")

diff_lim <- max(abs(c(raster_quantile(diff_r, 0.05),
                      raster_quantile(diff_r, 0.95))))
plot_map_conus(diff_r,
               colorscale = colorscale_diverging(
                 name = "Predicted -\nRAP (% cover)",
                 limits = c(-diff_lim, diff_lim)),
               title = "Predicted - RAP tree cover")
```

## Future climate

Predicted cover under each future scenario, and the change from current.

```{r map-future, fig.width = 9, fig.height = 6}
walk(setdiff(scenarios, "current"), \(scenario) {
  change <- pred[[scenario]] - pred$current
  change_lim <- max(abs(c(raster_quantile(change, 0.01),
                          raster_quantile(change, 0.99))))

  print(plot_map_conus(pred[[scenario]], colorscale = cover_scale,
                       title = paste("Predicted", str_to_lower(cover_label),
                                     scenario)))

  print(plot_map_conus(change,
                       colorscale = colorscale_diverging(
                         name = "Change\n(% cover)",
                         limits = c(-change_lim, change_lim)),
                       title = paste("Change in cover,", scenario,
                                     "vs current")))
})
```

# Session info

```{r session-info}
sessionInfo()
```
