---
title: "Classification model diagnostics"
author: "Martin Holdrege"
date: "`r Sys.Date()`"
params:
  cover_type: "classification"
  cover_model: null   # null uses opt$cover_model
  vc: null            # null uses opt$vc
  vmc: null           # null uses opt$vmc
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# Setup

```{r setup-data}
source("Functions/init.R")
source_functions()
library(glmnet)
theme_set(theme_custom1())

cover_type  <- params$cover_type
cover_model <- if (is.null(params$cover_model)) opt$cover_model else params$cover_model
vc          <- if (is.null(params$vc)) opt$vc else params$vc
vmc         <- if (is.null(params$vmc)) opt$vmc else params$vmc

fit  <- read_cover_model(cover_type, cover_model, vc, vmc)
if(is.null(fit$config$spec$engine)) fit$config$spec$engine <- 'glmnet' # for legacy reasons
spec <- fit$config$spec

# "glmnet" or "ranger"; the fit summary and a couple of labels differ

engine <- spec$engine
# for ranger, pred_in holds out-of-bag rather than in-sample predictions
in_label <- if (engine == "ranger") "out_of_bag" else "in_sample"
in_title <- if (engine == "ranger") {
  "Out-of-bag"
} else {
  "In-sample (model fit to all data)"
}

# predictors in original units (log1p_MAP is plotted as MAP)
source_vars <- unique(str_remove(spec$pred_vars, "^log1p_"))

# training data joined to the saved predictions; observed class rebuilt from
# the spec
dat <- read_cover_training(vc) |>
  inner_join(fit$predictions, by = c("cell", "year")) |>
  mutate(observed = as.integer(.data[[spec$response]] > spec$cover_threshold))
stopifnot(nrow(dat) == nrow(fit$predictions))

scenarios <- c("current", "BNU-ESM", "IPSL-CM5A-MR")
pred_dir <- file.path(paths$large, "Data_processed", "CoverData", "Predictions")
```

# Model info 

Model `r cover_type`/`r cover_model`, data `r vc`, version `r vmc`, engine
`r engine`: `r nrow(dat)` pixel-years; cutoff `r round(fit$threshold, 3)`

```{r info}
print(spec)
```

# Fit

## Fitted model

For glmnet, the cross-validation curve (binomial deviance vs log(lambda);
dashed lines mark min and 1se) and the non-zero coefficients at the selected
lambda. For ranger, the forest settings and the out-of-bag error.
for these binomial models (where it's succeses out of 1 try) binomial deviance is just -2*log likelihood

```{r fit-summary}
if (engine == "glmnet") {
  
  plot(fit$fit)
  
coefs <- coef(fit$fit, s = fit$lambda)
tibble(term = rownames(coefs), coefficient = as.numeric(coefs)) |>
  filter(coefficient != 0) |>
  arrange(desc(abs(coefficient))) |>
    knitr::kable(digits = 3) |>
    print()
  
} else if (engine == "ranger") {
  
  # selected fields only: print(fit$fit) deparses fit$call, which can hold
  # the whole training matrix. prediction.error is the out-of-bag Brier score.
  # Add importance = "impurity" to spec$ranger for variable importance.
  fit$fit[c("treetype", "num.trees", "mtry", "min.node.size", "splitrule",
            "num.samples", "num.independent.variables",
            "prediction.error")] |>
    map_chr(\(v) format(v, digits = 3)) |>
    enframe(name = "setting", value = "value") |>
    knitr::kable() |>
    print()
  
}
```

## Classification performance

AUC and the confusion matrix at the cutoff. Out-of-fold predictions come from
the environmental folds; for glmnet those are the same folds that chose
lambda, so they are mildly optimistic. The other set is in-sample for glmnet
and out-of-bag for ranger.

```{r performance}
perf <- function(pred) {
  pa <- data.frame(id = seq_len(nrow(dat)), obs = dat$observed, pred = pred)
  cm <- PresenceAbsence::cmx(pa, threshold = fit$threshold) # confusion matrix
  tibble(
    #the probability that a randomly chosen forest pixel gets a higher p̂ than a
    #randomly chosen non-forest pixel. 0.5 is random and 1 is perfect ranking.
          AUC = PresenceAbsence::auc(pa, st.dev = FALSE),
         PCC = PresenceAbsence::pcc(cm, st.dev = FALSE), # percent correctly classified
         # Of the pixels that are actually forest, it's the fraction called forest.
         sensitivity = PresenceAbsence::sensitivity(cm, st.dev = FALSE),
         # Of the pixels that are actually non-forest, it's the fraction called non-forest.
         specificity = PresenceAbsence::specificity(cm, st.dev = FALSE),
         observed_prevalence = mean(dat$observed),
         predicted_prevalence = mean(pred > fit$threshold))
}

bind_rows(perf(dat$pred_in), perf(dat$pred_oof)) |>
  mutate(predictions = c(in_label, "out_of_fold"), .before = 1) |>
  knitr::kable(digits = 3)

# confusion matrix (counts of pixel-years), out-of-fold
table(observed = dat$observed,
      predicted = as.integer(dat$pred_oof > fit$threshold)) |>
  knitr::kable(caption = "Out-of-fold confusion matrix (0 = not forest, 1 = forest)")
```

# Quantile plots

Points: proportion of pixel-years that are forest, observed and predicted
(predicted class = probability above the cutoff), within quantiles of each
predictor. Both are classes, so they're directly comparable.

```{r quantile-data}
# decile_dotplot() expects the predicted column to be <yvar>_pred
quant_df <- function(pred_col) {
  dat |>
    mutate(forest = observed,
           forest_pred = as.integer(.data[[pred_col]] > fit$threshold)) |>
    predvars2deciles(response_vars = c("forest", "forest_pred"),
                     pred_vars = source_vars)
}
quant_in  <- quant_df("pred_in")
quant_oof <- quant_df("pred_oof")
```

```{r quantile-plots, fig.width = 10, fig.height = 7}
decile_dotplot("forest", quant_in, ylab = "Proportion forest",
               add_predicted = TRUE, title = in_title,
               add_ribbon = FALSE)
decile_dotplot("forest", quant_oof, ylab = "Proportion forest",
               add_predicted = TRUE, title = "Out-of-fold",
               add_ribbon = FALSE)
```

## Filtered quantile plots

Only pixel-years in the lower and upper 20% of each filter variable, to see
whether the fit holds at the climate extremes.

```{r quantile-filtered, fig.width = 10, fig.height = 9}
filtered <- function(pred_col, title) {
  dat |>
  mutate(forest = observed,
         forest_pred = as.integer(.data[[pred_col]] > fit$threshold)) |>
  predvars2deciles(response_vars = c("forest", "forest_pred"),
                   pred_vars = source_vars, filter_var = TRUE,
                   filter_vars = source_vars) |>
  decile_dotplot_filtered("forest", df = _, ylab = "Proportion forest",
                          title = title)
}

filtered('pred_in', title = paste0('filtered (', in_title, ')'))
filtered('pred_oof', title = 'filtered (out of fold)')

```

# Partial dependence

Mean predicted probability as each predictor varies, others at their
observed values.

```{r pdp, fig.width = 10, fig.height = 6}
pdp_classification(fit, dat) |>
  plot_pdp(ylab = "Predicted probability of forest")
```

# Accumulated local effects

Change in predicted probability as each predictor increases, relative to
the average prediction (Apley & Zhu 2020). Built from small steps within
quantile bins, using only pixels near each value, so unlike the partial
dependence plots it does not evaluate the model at combinations of
predictors that don't occur. Where the two disagree, the PDP is being
driven by those combinations. Middle 98% of each predictor. Points are
bin edges; each segment is the mean change in probability across a bin.

```{r ale, fig.width = 10, fig.height = 6}
ale_classification(fit, dat) |>
  plot_ale()
```

## Filtered accumulated local effects

ALE computed separately within the lowest and highest 20% of each filter
variable, as in the filtered quantile plots. Each line is the subset's mean
predicted probability plus its ALE, so the lines differ in level where the
subsets differ, and in shape where the predictor interacts with the filter
variable in the fitted model.

```{r ale-filtered, fig.width = 10, fig.height = 9}
ale_classification_filtered(fit, dat) |>
  plot_ale_filtered()
```

# Maps

## Misclassification at training pixels

Observed class vs predicted class, for both sets of predictions. One point
per pixel, using its most recent year.

```{r map-misclass, fig.width = 9, fig.height = 6}
plot_misclass <- function(pred_col, title) {
  misclass_sf <- dat |>
    slice_max(year, n = 1, by = cell) |>
    mutate(predicted = as.integer(.data[[pred_col]] > fit$threshold),
           outcome = case_when(
             observed == 1 & predicted == 1 ~ "forest, correct",
             observed == 0 & predicted == 0 ~ "not forest, correct",
             observed == 0 & predicted == 1 ~ "false forest",
             observed == 1 & predicted == 0 ~ "missed forest"
           )) |>
    sf::st_as_sf(coords = c("x", "y"), crs = terra::crs(read_mask()))
  
  plot_points_conus(misclass_sf, color_var = "outcome", point_size = 0.3,
                    title = title,
                    colorscale = scale_colour_manual(
                      name = NULL,
                      values = c("forest, correct" = "darkgreen",
                                 "not forest, correct" = "grey75",
                                 "false forest" = "#2166ac",
                                 "missed forest" = "orange"))) +
    # larger points in the legend only
    guides(colour = guide_legend(override.aes = list(size = 3)))
}

plot_misclass("pred_in", in_title)
plot_misclass("pred_oof", "Out-of-fold")
```

## Out-of-fold residuals at training pixels

Mean over years of (predicted probability - observed class) per pixel.
Positive: predicted more forest than observed.

```{r map-residuals, fig.width = 9, fig.height = 6}
resid_sf <- dat |>
  summarise(resid = mean(pred_oof - observed), .by = c(cell, x, y)) |>
  sf::st_as_sf(coords = c("x", "y"), crs = terra::crs(read_mask()))

plot_points_conus(resid_sf, color_var = "resid", point_size = 0.3,
                  colorscale = scale_colour_gradient2(
                    name = "Predicted -\nobserved", low = "#b2182b",
                    high = "#2166ac", limits = c(-1, 1)))
```

## Current climate

All of CONUS (snap mask), not only the available area: predictions stand for
potential natural vegetation, and SOILWAT2 will be run CONUS-wide.

```{r map-read}
read_pred <- function(scenario) {
  terra::rast(file.path(pred_dir, paste0(cover_type, "_", cover_model, "_",
                                         vc, "-", vmc, "_", scenario,
                                         ".tif")))
}
pred <- map(scenarios, read_pred) |> set_names(scenarios)

# binary and change rasters as categories, so legends show classes rather
# than a continuous color bar
as_class_raster <- function(r) {
  r <- terra::as.int(r)
  levels(r) <- data.frame(value = c(0, 1), class = c("not forest", "forest"))
  r
}
class_scale <- scale_fill_manual(name = NULL,
                                 values = c("not forest" = "grey85",
                                            "forest" = "darkgreen"),
                                 na.translate = FALSE)

as_change_raster <- function(r) {
  r <- terra::as.int(r)
  levels(r) <- data.frame(value = c(-1, 0, 1),
                          change = c("forest lost", "no change",
                                     "forest gained"))
  r
}
change_scale <- scale_fill_manual(name = NULL,
                                  values = c("forest lost" = "orange",
                                             "no change" = "grey90",
                                             "forest gained" = "#1b7837"),
                                  na.translate = FALSE)
```

```{r raster-summary}
terra::global(pred$current$prob, c("min", "mean", "max"), na.rm = TRUE)
```



```{r map-current, fig.width = 9, fig.height = 6}
plot_map_conus(pred$current$prob,
               colorscale = scale_fill_gradient2(
                    name = "Probability", 
                    low = "#b2182b",
                    high = "#2166ac",
                    mid = 'grey85',
                    midpoint = fit$threshold,
                    na.value = 'transparent',
                    limits = c(0, 1)),
               title = "Predicted probability of forest")

plot_map_conus(as_class_raster(pred$current$class),
               colorscale = class_scale,
               title = "Predicted class")
```

## Future climate

Class under each future scenario, and the change from current (+1 becomes
forest, -1 stops being forest).

```{r map-future, fig.width = 9, fig.height = 6}
walk(setdiff(scenarios, "current"), \(scenario) {
  change <- pred[[scenario]]$class - pred$current$class
  
  print(plot_map_conus(as_class_raster(pred[[scenario]]$class),
                       colorscale = class_scale,
                       title = paste("Predicted class,", scenario)))
  
  print(plot_map_conus(as_change_raster(change),
                       colorscale = change_scale,
                       title = paste("Change in class,", scenario,
                                     "vs current")))
})
```

# Session info

```{r session-info}
sessionInfo()
```