---
title: "Combined tree cover diagnostics"
author: "Martin Holdrege"
date: "`r Sys.Date()`"
params:
  vmc: null   # combination version; null uses opt$vmc
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# Setup

```{r setup-data}
source("Functions/init.R")
source_functions()
theme_set(theme_custom1())

vmc <- if (is.null(params$vmc)) opt$vmc else params$vmc

# component models and blending parameters (04_combine_tree.R)
blend <- read_csv(combined_tree_path(vmc, "blend"), show_col_types = FALSE)

scenarios <- c("current", "BNU-ESM", "IPSL-CM5A-MR")
pred <- map(scenarios, \(s) terra::rast(combined_tree_path(vmc, "prediction", s))) |>
  set_names(scenarios)

# observed: 2021 RAP tree cover, set to 0 where the RAP zero-tree layer says
# zero tree (with the threshold the zero-tree model was fit with). Cells
# missing either layer are NA
zt <- filter(blend, component == "zero_tree")
zt_threshold <- read_cover_model(zt$cover_type, zt$cover_model, zt$vc,
                                 zt$vmc)$config$spec$cover_threshold

rap_tree <- terra::rast(file.path(paths$large, "Data_processed/CoverData/rap",
                                  "RAP_v3_cover_2021_1000m.tif"))[["tree"]]
rap_zero <- 100 * read_zero_tree_raster("fracZeroTree") > zt_threshold
# GEE exports are one cell larger than the snap grid
rap_tree <- align_raster(rap_tree, pred$current)
rap_zero <- align_raster(rap_zero, pred$current)
obs <- terra::ifel(rap_zero == 1, 0, rap_tree) |>
  terra::mask(rap_tree) |>  # NA where either layer is missing
  terra::mask(rap_zero)
names(obs) <- "observed"
stopifnot(terra::compareGeom(obs, pred$current))

pred_cur <- terra::mask(pred$current$cover, obs)  # cells with RAP data
diff_r <- pred_cur - obs


# colour scale for cover: exact zeros get their own colour
upper <- raster_quantile(obs, 0.99)
cover_scale <- colorscale_cover(name = "Tree\ncover (%)", limits = c(0, upper),
                                zero_color = "tan")
```

# Combination

Component models and blending parameters. Away from the forest boundary the
forest or non-forest model is used as is; between `score_low` and
`score_high` (`blend_prop` of CONUS cells on each side of the forest
threshold, under current climate) the two are blended, with the weight on
the forest model rising linearly from 0 to 1 through 0.5 at the threshold.
After blending, tree cover is set to 0 in cells classified non-forest and
zero tree.

```{r blend-table}
knitr::kable(blend, digits = 3)
```

```{r map-weight, fig.width = 9, fig.height = 6}

# weight 0 (non-forest model only) and 1 (forest model only) get their own
# colours; the blending band (0 < w < 1) gets a separate ramp
eps  <- 1e-6
ramp <- viridisLite::plasma(9, begin = 0.1, end = 0.9)
weight_scale <- scale_fill_gradientn(
  name = "Weight on\nforest model",
  colours = c("grey85", ramp, "darkgreen"),
  values  = c(0, seq(eps, 1 - eps, length.out = length(ramp)), 1),
  limits  = c(0, 1),
  na.value = "transparent"
)

plot_map_conus(pred$current$weight, colorscale = weight_scale,
               title = "Blending weight (current climate)") +
  labs(subtitle = "Grey: non-forest model only (weight 0). Green: forest model only (weight 1).")
```

# Current climate vs RAP

Observed is 2021 RAP tree cover, with zero-tree cells (RAP zero-tree layer)
set to 0, in the cells passing the LCMAP and fire masks. Predictions are
compared in those cells only. Exact zeros are shown in tan, so they can be
told apart from low cover. The training data were sampled from these rasters,
so this shows the spatial pattern of the errors rather than an independent
test. Metrics are over all cells with both layers.

```{r metrics}
v <- terra::values(c(obs, pred_cur), mat = TRUE, na.rm = TRUE)
cover_metrics(v[, "observed"], v[, "cover"]) |>
  knitr::kable(digits = 3)
rm(v)
```

```{r map-current, fig.width = 9, fig.height = 6}
plot_map_conus(obs, colorscale = cover_scale,
               title = "RAP tree cover, 2021 (zero-tree cells set to 0)")
plot_map_conus(pred_cur, colorscale = cover_scale,
               title = "Predicted tree cover (same cells)")
plot_map_conus(pred$current$cover, colorscale = cover_scale,
               title = "Predicted tree cover (all cells)")

diff_lim <- max(abs(c(raster_quantile(diff_r, 0.05),
                      raster_quantile(diff_r, 0.95))))
plot_map_conus(diff_r,
               colorscale = colorscale_diverging(
                 name = "Predicted -\nRAP (% cover)",
                 limits = c(-diff_lim, diff_lim)),
               title = "Predicted - RAP tree cover",
               inset_hist = TRUE)
```

# Future climate

Predicted tree cover under each future scenario, and the change from current
(all cells).

```{r map-future, fig.width = 9, fig.height = 6}
walk(setdiff(scenarios, "current"), \(scenario) {
  change <- pred[[scenario]]$cover - pred$current$cover
  change_lim <- max(abs(c(raster_quantile(change, 0.01),
                          raster_quantile(change, 0.99))))

  g1 <- plot_map_conus(pred[[scenario]]$cover, colorscale = cover_scale,
                       title = paste("Predicted tree cover,", scenario))

  g2 <- plot_map_conus(change,
                       colorscale = colorscale_diverging(
                         name = "Change\n(% cover)",
                         limits = c(-change_lim, change_lim)),
                       title = paste("Change in tree cover,", scenario,
                                     "vs current"),
        inset_hist = TRUE)


})
```

# Session info

```{r session-info}
sessionInfo()
```
